# Inverse Probability of Treatment Weighting (IPTW) {#chap_iptw}

## Estimation of the Average total effect

### IPTW for the ATE
If the average total effect (ATE) is identifiable, $\Psi_{ATE} = \mathbb{E}(Y_{A=1}) - \mathbb{E}(Y_{A=0})$ can be expressed using Inverse probability of treatment weighting (IPTW), denoting $\mathbb{P}(A=a \mid L(0)) = g(A=a \mid L(0))$:
\begin{equation}
\Psi_{ATE} = \mathbb{E}\left( \frac{\mathbb{I}(A=1)}{g(A=1 \mid L(0))} Y \right) - \mathbb{E}\left( \frac{\mathbb{I}(A=0)}{g(A=0 \mid L(0))} Y \right)
\end{equation}
  
The following steps describe the implementation of the IPTW estimator


1. Estimate the treatment mechanism $g(A=1 \mid L(0))$

2. Predict each individual's probability of being exposed to her own exposure

3. Apply weights corresponding to the inverse of the predicted probability $w_i = \frac{1}{\hat{g}(A_i = a \mid L(0)_i)}$

4. Use the empirical mean of the weighted outcome $Y$: $\hat{\mathbb{E}}(Y_a) = \frac{1}{n} \sum_{i=1}^n \frac{\mathbb{I}(A_i=a)}{\hat{g}(A_i=a \mid L(0)_i)} Y_i$

```{r Psi_ATE_iptw, echo=TRUE, eval = FALSE}
# 1. Estimate g
g.L <- glm(A0_ace ~ L0_male + L0_parent_low_educ_lv, 
           family = "binomial", data = df2_int)

# 2. Predict each individual's probability of being exposed to her own exposure
# predict the probabilities P(A0_ace=1|L(0)) & P(A0_ace=0|L(0))
pred.g1.L <- predict(g.L, type="response")
pred.g0.L <- 1 - pred.g1.L
# the predicted probability of the observed treatment A_i=a is :
gA.L <- rep(NA, nrow(df2_int))
gA.L[df2_int$A0_ace == 1] <- pred.g1.L[df2_int$A0_ace == 1]
gA.L[df2_int$A0_ace == 0] <- pred.g0.L[df2_int$A0_ace == 0]

# 3. Apply weights corresponding to the inverse of the predicted probability
wt <- 1 / gA.L

# 4. Use the empirical mean of the weighted outcome
# point estimates:
IPTW.death <- mean(wt * as.numeric(df2_int$A0_ace == 1) * df2_int$Y_death) -
  mean(wt * as.numeric(df2_int$A0_ace == 0) * df2_int$Y_death)
IPTW.death
# [1] 0.08224947

IPTW.qol <- mean(wt * as.numeric(df2_int$A0_ace == 1) * df2_int$Y_qol) -
  mean(wt * as.numeric(df2_int$A0_ace == 0) * df2_int$Y_qol)
IPTW.qol
# [1] -8.436797
```

The ATE estimates using IPTW for death probability and mean quality of life are respectively +8.2\% and -8.44.

### Stabilized IPTW for the ATE
If the average total effect (ATE) is identifiable, $\Psi_{ATE}$ can be estimated using a stabilized IPTW estimator:
\begin{equation}
\hat{\mathbb{E}}(Y_1) - \hat{\mathbb{E}}(Y_0) =  \frac{\frac{1}{n} \sum_{i=1}^n \frac{\mathbb{I}(A_i=1)\hat{g}^*(A_i=1)}{\hat{g}(A_i=1 \mid L(0)_i)} Y_i}{ \frac{1}{n} \sum_{i=1}^n \frac{\mathbb{I}(A_i=1)\hat{g}^*(A_i=1)}{\hat{g}(A_i=1 \mid L(0)_i)}} - \frac{\frac{1}{n} \sum_{i=1}^n \frac{\mathbb{I}(A_i=0)\hat{g}^*(A_i=0)}{\hat{g}(A_i=0 \mid L(0)_i)} Y_i}{ \frac{1}{n} \sum_{i=1}^n \frac{\mathbb{I}(A_i=0)\hat{g}^*(A_i=0)}{\hat{g}(A_i=0 \mid L(0)_i)}}
\end{equation}
The estimation algorithm is the same as for IPTW, but taking into account any non-null function of $A$ ($g^*(A_i=a)$) in the denominator of the weight in step 3, and applying the stabilized estimator in step 4.

```{r Psi_ATE_siptw, echo=TRUE, eval = FALSE}
# 3. For example, applying g^*(A) = 1
# 4. Applying the stabilized estimator
# point estimates:
s.IPTW.death <- (mean(wt * as.numeric(df2_int$A0_ace == 1) * df2_int$Y_death) /
                   mean(wt * as.numeric(df2_int$A0_ace == 1))) -
  (mean(wt * as.numeric(df2_int$A0_ace == 0) * df2_int$Y_death) /
     mean(wt * as.numeric(df2_int$A0_ace == 0)))
s.IPTW.death
# [1] 0.08294185

s.IPTW.qol <- (mean(wt * as.numeric(df2_int$A0_ace == 1) * df2_int$Y_qol) /
                 mean(wt * as.numeric(df2_int$A0_ace == 1))) -
  (mean(wt * as.numeric(df2_int$A0_ace == 0) * df2_int$Y_qol) /
     mean(wt * as.numeric(df2_int$A0_ace == 0)))
s.IPTW.qol
# [1] -8.291992
```

The ATE estimates using stabilized IPTW for death probability and mean quality of life are respectively +8.3\% and -8.29.


## Estimation of the Controlled direct effect (CDE)

### IPTW for the CDE
If the controlled direct effect (CDE) is identifiable, $\Psi^{\text{CDE}_m} = \mathbb{E}(Y_{A=1,M=m}) - \mathbb{E}(Y_{A=0,M=m})$ can be expressed by the basic Horvitz Thompson estimator (using Inverse probability of treatment weighting (IPTW)), denoting $\mathbb{P}\left(A=a \mid L(0)) = g(A=a \mid L(0)\right)$ and $\mathbb{P}\left(M=m \mid L(0),A,L(1)) = g(M=m \mid L(0),A,L(1)\right)$:
\begin{equation}
\small
\Psi^{\text{CDE}_m} = \mathbb{E}\left[ \frac{\mathbb{I}(A=1 \cap M=m)}{g(A=1 \mid L(0)) \times g(M=m \mid L(0),A,L(1))} Y \right] - \mathbb{E}\left[ \frac{\mathbb{I}(A=0 \cap M=m)}{g(A=0 \mid L(0)) \times g(M=m \mid L(0),A,L(1))} Y \right]
\end{equation}
  
The following steps describe the implementation of the IPTW estimator

1. Estimate the treatment mechanisms $g\left(A=1 \mid L(0)\right)$ and $g\left(M=1 \mid L(0),A,L(1)\right)$

2. Predict each individual's probability of being exposed to her own exposure

3. Apply weights corresponding to the inverse of the predicted probability $w_{A_i} = \frac{1}{\hat{g}(A_i = a \mid L(0)_i)}$ and $w_{M_i} = \frac{1}{\hat{g}(M_i = m \mid L(0)_i,A_i,L(1)_i)}$

4. Use the empirical mean of the weighted outcome $Y$: $\hat{\mathbb{E}}(Y_{a,m}) = \frac{1}{n} \sum_{i=1}^n \frac{\mathbb{I}(A_i=a \cap M_i=m)}{\hat{g}(A_i=a \mid L(0)_i) \times \hat{g}(M_i=m \mid L(0)_i,A_i,L(1)_i)} Y_i$

```{r Psi_CDE_iptw, echo=TRUE, eval = FALSE}
# 1. Estimate gA and gM
gA.L <- glm(A0_ace ~ L0_male + L0_parent_low_educ_lv, 
            family = "binomial", data = df2_int)
gM.L <- glm(M_smoking ~ L0_male + L0_parent_low_educ_lv + A0_ace + L1, 
            family = "binomial", data = df2_int)

# 2. Predict each individual's probability of being exposed to her own exposure
# predict the probabilities P(A0_ace=1|L(0)) & P(A0_ace=0|L(0))
pred.gA1.L <- predict(gA.L, type = "response")
pred.gA0.L <- 1 - pred.gA1.L
# the predicted probability of the observed treatment A_i=a is :
gAobs.L <- rep(NA, nrow(df2_int))
gAobs.L[df2_int$A0_ace == 1] <- pred.gA1.L[df2_int$A0_ace == 1]
gAobs.L[df2_int$A0_ace == 0] <- pred.gA0.L[df2_int$A0_ace == 0]

# predict the probabilities P(M=1|L(0),A,L(1)) & P(M=0|L(0),A,L(1))
pred.gM1.L <- predict(gM.L, type = "response")
pred.gM0.L <- 1 - pred.gM1.L
# the predicted probability of the observed treatment M_i=m is :
gMobs.L <- rep(NA, nrow(df2_int))
gMobs.L[df2_int$M_smoking == 1] <- pred.gM1.L[df2_int$M_smoking == 1]
gMobs.L[df2_int$M_smoking == 0] <- pred.gM0.L[df2_int$M_smoking == 0]

# 3. Apply weights corresponding to the inverse of the predicted probability
wt_A <- 1 / gAobs.L
wt_M <- 1 / gMobs.L
wt <- wt_A * wt_M

# 4. Use the empirical mean of the weighted outcome
# point estimates of CDE, setting M=0
CDE_IPTW_m0_death <- (mean(wt * as.numeric(df2_int$A0_ace == 1 & 
                                             df2_int$M_smoking == 0) * 
                             df2_int$Y_death) -
                        mean(wt * as.numeric(df2_int$A0_ace==0 & 
                                               df2_int$M_smoking == 0) * 
                               df2_int$Y_death))
CDE_IPTW_m0_death
# [1] 0.05874684

CDE_IPTW_m0_qol <- (mean(wt * as.numeric(df2_int$A0_ace == 1 & 
                                           df2_int$M_smoking == 0) * 
                           df2_int$Y_qol) -
                      mean(wt * as.numeric(df2_int$A0_ace==0 & 
                                             df2_int$M_smoking == 0) * 
                             df2_int$Y_qol))
CDE_IPTW_m0_qol
# [1] -5.341138

# point estimates of CDE, setting M=1
CDE_IPTW_m1_death <- (mean(wt * as.numeric(df2_int$A0_ace == 1 & 
                                             df2_int$M_smoking == 1) * 
                             df2_int$Y_death) -
                        mean(wt * as.numeric(df2_int$A0_ace==0 & 
                                               df2_int$M_smoking == 1) * 
                               df2_int$Y_death))
CDE_IPTW_m1_death
# [1] 0.101733

CDE_IPTW_m1_qol <- (mean(wt * as.numeric(df2_int$A0_ace == 1 & 
                                           df2_int$M_smoking == 1) * 
                           df2_int$Y_qol) -
                      mean(wt * as.numeric(df2_int$A0_ace==0 & 
                                             df2_int$M_smoking == 1) * 
                             df2_int$Y_qol))
CDE_IPTW_m1_qol
# [1] -8.185866
```

### Stabilized IPTW for the CDE
If the controlled dired effect (CDE) is identifiable, $\Psi^{CDE}$ can be estimated using a stabilized IPTW estimator (modified Horvitz Thompson estimator):
\begin{equation}
\small
\hat{\mathbb{E}}(Y_{1,m}) - \hat{\mathbb{E}}(Y_{0,m}) =  \frac{ \sum_{i=1}^n \frac{\mathbb{I}(A_i=1 \cap M_i=m)}{\hat{g}(A_i=1 \mid L(0)_i) \times \hat{g}(M_i=m \mid L(0)_i,A_i,L(1)_i)} Y_i}{\sum_{i=1}^n \frac{\mathbb{I}(A_i=1 \cap M_i=m)}{\hat{g}(A_i=1 \mid L(0)_i) \hat{g}(M_i=m \mid L(0)_i,A_i,L(1)_i)}} - \frac{ \sum_{i=1}^n \frac{\mathbb{I}(A_i=0 \cap M_i=m)}{\hat{g}(A_i=0 \mid L(0)_i) \times \hat{g}(M_i=m \mid L(0)_i,A_i,L(1)_i)} Y_i}{ \sum_{i=1}^n \frac{\mathbb{I}(A_i=0 \cap M_i=m)}{\hat{g}(A_i=0 \mid L(0)_i) \times \hat{g}(M_i=m \mid L(0)_i,A_i,L(1)_i)}}
\end{equation}
The estimation algorithm is the same as for IPTW, but applying the stabilized estimator in step 4.

```{r Psi_CDE_siptw, echo=TRUE, eval = FALSE}
# 4. Applying the stabilized estimator
# point estimates of CDE, setting M=0:
CDE_sIPTW_m0_death <- (mean(wt * as.numeric(df2_int$A0_ace == 1 & 
                                              df2_int$M_smoking == 0) * 
                              df2_int$Y_death) / 
                         mean(wt * as.numeric(df2_int$A0_ace == 1 & 
                                                df2_int$M_smoking == 0))) -
                         (mean(wt * as.numeric(df2_int$A0_ace == 0 & 
                                              df2_int$M_smoking == 0) * 
                              df2_int$Y_death) / 
                         mean(wt * as.numeric(df2_int$A0_ace == 0 & 
                                                df2_int$M_smoking == 0)))
CDE_sIPTW_m0_death
# [1] 0.0601292

CDE_sIPTW_m0_qol <- (mean(wt * as.numeric(df2_int$A0_ace == 1 & 
                                              df2_int$M_smoking == 0) * 
                              df2_int$Y_qol) / 
                         mean(wt * as.numeric(df2_int$A0_ace == 1 & 
                                                df2_int$M_smoking == 0))) -
                         (mean(wt * as.numeric(df2_int$A0_ace == 0 & 
                                              df2_int$M_smoking == 0) * 
                              df2_int$Y_qol) / 
                         mean(wt * as.numeric(df2_int$A0_ace == 0 & 
                                                df2_int$M_smoking == 0)))
CDE_sIPTW_m0_qol
# [1] -4.966328

# point estimates of CDE, setting M=1:
CDE_sIPTW_m1_death <- (mean(wt * as.numeric(df2_int$A0_ace == 1 & 
                                              df2_int$M_smoking == 1) * 
                              df2_int$Y_death) / 
                         mean(wt * as.numeric(df2_int$A0_ace == 1 & 
                                                df2_int$M_smoking == 1))) -
                         (mean(wt * as.numeric(df2_int$A0_ace == 0 & 
                                              df2_int$M_smoking == 1) * 
                              df2_int$Y_death) / 
                         mean(wt * as.numeric(df2_int$A0_ace == 0 & 
                                                df2_int$M_smoking == 1)))
CDE_sIPTW_m1_death
# [1] 0.09030186

CDE_sIPTW_m1_qol <- (mean(wt * as.numeric(df2_int$A0_ace == 1 & 
                                              df2_int$M_smoking == 1) * 
                              df2_int$Y_qol) / 
                         mean(wt * as.numeric(df2_int$A0_ace == 1 & 
                                                df2_int$M_smoking == 1))) -
                         (mean(wt * as.numeric(df2_int$A0_ace == 0 & 
                                              df2_int$M_smoking == 1) * 
                              df2_int$Y_qol) / 
                         mean(wt * as.numeric(df2_int$A0_ace == 0 & 
                                                df2_int$M_smoking == 1)))
CDE_sIPTW_m1_qol
# [1] -10.03045
```
