# Inverse Probability of Treatment Weighting (IPTW) {#ChapIptw}

## Estimation of the Average total effect

### IPTW for the ATE
If the average total effect (ATE) is identifiable, $\Psi_{ATE} = \mathbb{E}(Y_{A=1}) - \mathbb{E}(Y_{A=0})$ can be expressed using Inverse probability of treatment weighting (IPTW), denoting $\mathbb{P}(A=a \mid L(0)) = g(A=a \mid L(0))$:
\begin{equation}
\Psi_{ATE} = \mathbb{E}\left( \frac{\mathbb{I}(A=1)}{g(A=1 \mid L(0))} Y \right) - \mathbb{E}\left( \frac{\mathbb{I}(A=0)}{g(A=0 \mid L(0))} Y \right)
\end{equation}
  
The following steps describe the implementation of the IPTW estimator


1. Estimate the treatment mechanism $g(A=1 \mid L(0))$

2. Predict each individual's probability of being exposed to her own exposure

3. Apply weights corresponding to the inverse of the predicted probability $w_i = \frac{1}{\hat{g}(A = a_i \mid L(0)_i)}$

4. Use the empirical mean of the weighted outcome $Y$: $\hat{\mathbb{E}}(Y_a) = \frac{1}{n} \sum_{i=1}^n \frac{\mathbb{I}(A_i=a)}{\hat{g}(A=a_i \mid L(0)_i)} Y_i$

```{r Psi_ATE_iptw, echo=TRUE, eval = FALSE}
## 1. Estimate g
g.L <- glm(A0_PM2.5 ~ L0_male + L0_soc_env, 
           family = "binomial", data = df2_int)

## 2. Predict each individual's probability of being exposed to her own exposure
# predict the probabilities P(A0_PM2.5=1|L(0)) & P(A0_PM2.5=0|L(0))
pred.g1.L <- predict(g.L, type="response")
pred.g0.L <- 1 - pred.g1.L
# the predicted probability of the observed treatment A=a_i is :
gA.L <- rep(NA, nrow(df2_int))
gA.L[df2_int$A0_PM2.5 == 1] <- pred.g1.L[df2_int$A0_PM2.5 == 1]
gA.L[df2_int$A0_PM2.5 == 0] <- pred.g0.L[df2_int$A0_PM2.5 == 0]

## 3. Apply weights corresponding to the inverse of the predicted probability
wt <- 1 / gA.L

## 4. Use the empirical mean of the weighted outcome
# point estimates:
IPTW.death <- mean(wt * as.numeric(df2_int$A0_PM2.5 == 1) * df2_int$Y_death) -
  mean(wt * as.numeric(df2_int$A0_PM2.5 == 0) * df2_int$Y_death)
IPTW.death
# [1] 0.08224947

IPTW.qol <- mean(wt * as.numeric(df2_int$A0_PM2.5 == 1) * df2_int$Y_qol) -
  mean(wt * as.numeric(df2_int$A0_PM2.5 == 0) * df2_int$Y_qol)
IPTW.qol
# [1] -8.436797
```

The ATE estimates using IPTW for death probability and mean quality of life are respectively +8.2\% and -8.44.

### Stabilized IPTW for the ATE
If the average total effect (ATE) is identifiable, $\Psi_{ATE}$ can be estimated using a stabilized IPTW estimator:
\begin{equation}
\hat{\mathbb{E}}(Y_1) - \hat{\mathbb{E}}(Y_0) =  \frac{\frac{1}{n} \sum_{i=1}^n \frac{\mathbb{I}(A_i=1)\hat{g}^*(A_i=1)}{\hat{g}(A_i=1 \mid L(0)_i)} Y_i}{ \frac{1}{n} \sum_{i=1}^n \frac{\mathbb{I}(A_i=1)\hat{g}^*(A_i=1)}{\hat{g}(A_i=1 \mid L(0)_i)}} - \frac{\frac{1}{n} \sum_{i=1}^n \frac{\mathbb{I}(A_i=0)\hat{g}^*(A_i=0)}{\hat{g}(A_i=0 \mid L(0)_i)} Y_i}{ \frac{1}{n} \sum_{i=1}^n \frac{\mathbb{I}(A_i=0)\hat{g}^*(A_i=0)}{\hat{g}(A_i=0 \mid L(0)_i)}}
\end{equation}
The estimation algorithm is the same as for IPTW, but taking into account any non-null function of $A$ ($g^*(A_i=a)$) in the denominator of the weight in step 3, and applying the stabilized estimator in step 4.

```{r Psi_ATE_siptw, echo=TRUE, eval = FALSE}
## 3. For example, applying g^*(A) = 1
## 4. Applying the stabilized estimator
# point estimates:
s.IPTW.death <- (mean(wt * as.numeric(df2_int$A0_PM2.5 == 1) * df2_int$Y_death) /
                   mean(wt * as.numeric(df2_int$A0_PM2.5 == 1))) -
  (mean(wt * as.numeric(df2_int$A0_PM2.5 == 0) * df2_int$Y_death) /
     mean(wt * as.numeric(df2_int$A0_PM2.5 == 0)))
s.IPTW.death
# [1] 0.08294185

s.IPTW.qol <- (mean(wt * as.numeric(df2_int$A0_PM2.5 == 1) * df2_int$Y_qol) /
                 mean(wt * as.numeric(df2_int$A0_PM2.5 == 1))) -
  (mean(wt * as.numeric(df2_int$A0_PM2.5 == 0) * df2_int$Y_qol) /
     mean(wt * as.numeric(df2_int$A0_PM2.5 == 0)))
s.IPTW.qol
# [1] -8.291992
```

The ATE estimates using stabilized IPTW for death probability and mean quality of life are respectively +8.3\% and -8.29.


## Estimation of the Controlled direct effect (CDE)

### IPTW for the CDE
If the controlled direct effect (CDE) is identifiable, $\Psi^{\text{CDE}_m} = \mathbb{E}(Y_{A=1,M=m}) - \mathbb{E}(Y_{A=0,M=m})$ can be expressed by the basic Horvitz Thompson estimator (using Inverse probability of treatment weighting (IPTW)), denoting $\mathbb{P}\left(A=a \mid L(0)) = g(A=a \mid L(0)\right)$ and $\mathbb{P}\left(M=m \mid L(0),A,L(1)) = g(M=m \mid L(0),A,L(1)\right)$:
\begin{equation}
\small
\Psi^{\text{CDE}_m} = \mathbb{E}\left[ \frac{\mathbb{I}(A=1 \cap M=m)}{g(A=1 \mid L(0)) \times g(M=m \mid L(0),A,L(1))} Y \right] - \mathbb{E}\left[ \frac{\mathbb{I}(A=0 \cap M=m)}{g(A=0 \mid L(0)) \times g(M=m \mid L(0),A,L(1))} Y \right]
\end{equation}
  
The following steps describe the implementation of the IPTW estimator

1. Estimate the treatment mechanisms $g\left(A=1 \mid L(0)\right)$ and $g\left(M=1 \mid L(0),A,L(1)\right)$

2. Predict each individual's probability of being exposed to her own exposure

3. Apply weights corresponding to the inverse of the predicted probability $w_{A_i} = \frac{1}{\hat{g}(A = a_i \mid L(0)_i)}$ and $w_{M_i} = \frac{1}{\hat{g}(M = m_i \mid L(0)_i,A_i,L(1)_i)}$

4. Use the empirical mean of the weighted outcome $Y$: $\hat{\mathbb{E}}(Y_{a,m}) = \frac{1}{n} \sum_{i=1}^n \frac{\mathbb{I}(A_i=a \cap M_i=m)}{\hat{g}(A=a_i \mid L(0)_i) \times \hat{g}(M=m_i \mid L(0)_i,A_i,L(1)_i)} Y_i$

```{r Psi_CDE_iptw, echo=TRUE, eval = FALSE}
## 1. Estimate gA and gM
gA.L <- glm(A0_PM2.5 ~ L0_male + L0_soc_env, 
            family = "binomial", data = df2_int)
gM.L <- glm(M_diabetes ~ L0_male + L0_soc_env + A0_PM2.5 + L1, 
            family = "binomial", data = df2_int)

## 2. Predict each individual's probability of being exposed to her own exposure
# predict the probabilities P(A0_PM2.5=1|L(0)) & P(A0_PM2.5=0|L(0))
pred.gA1.L <- predict(gA.L, type = "response")
pred.gA0.L <- 1 - pred.gA1.L
# the predicted probability of the observed treatment A_i=a is :
gAobs.L <- rep(NA, nrow(df2_int))
gAobs.L[df2_int$A0_PM2.5 == 1] <- pred.gA1.L[df2_int$A0_PM2.5 == 1]
gAobs.L[df2_int$A0_PM2.5 == 0] <- pred.gA0.L[df2_int$A0_PM2.5 == 0]

# predict the probabilities P(M=1|L(0),A,L(1)) & P(M=0|L(0),A,L(1))
pred.gM1.L <- predict(gM.L, type = "response")
pred.gM0.L <- 1 - pred.gM1.L
# the predicted probability of the observed treatment M_i=m is :
gMobs.L <- rep(NA, nrow(df2_int))
gMobs.L[df2_int$M_diabetes == 1] <- pred.gM1.L[df2_int$M_diabetes == 1]
gMobs.L[df2_int$M_diabetes == 0] <- pred.gM0.L[df2_int$M_diabetes == 0]

## 3. Apply weights corresponding to the inverse of the predicted probability
wt_A <- 1 / gAobs.L
wt_M <- 1 / gMobs.L
wt <- wt_A * wt_M

## 4. Use the empirical mean of the weighted outcome
# point estimates of CDE, setting M=0
CDE_IPTW_m0_death <- (mean(wt * as.numeric(df2_int$A0_PM2.5 == 1 & 
                                             df2_int$M_diabetes == 0) * 
                             df2_int$Y_death) -
                        mean(wt * as.numeric(df2_int$A0_PM2.5==0 & 
                                               df2_int$M_diabetes == 0) * 
                               df2_int$Y_death))
CDE_IPTW_m0_death
# [1] 0.05874684

CDE_IPTW_m0_qol <- (mean(wt * as.numeric(df2_int$A0_PM2.5 == 1 & 
                                           df2_int$M_diabetes == 0) * 
                           df2_int$Y_qol) -
                      mean(wt * as.numeric(df2_int$A0_PM2.5==0 & 
                                             df2_int$M_diabetes == 0) * 
                             df2_int$Y_qol))
CDE_IPTW_m0_qol
# [1] -5.341138

# point estimates of CDE, setting M=1
CDE_IPTW_m1_death <- (mean(wt * as.numeric(df2_int$A0_PM2.5 == 1 & 
                                             df2_int$M_diabetes == 1) * 
                             df2_int$Y_death) -
                        mean(wt * as.numeric(df2_int$A0_PM2.5==0 & 
                                               df2_int$M_diabetes == 1) * 
                               df2_int$Y_death))
CDE_IPTW_m1_death
# [1] 0.101733

CDE_IPTW_m1_qol <- (mean(wt * as.numeric(df2_int$A0_PM2.5 == 1 & 
                                           df2_int$M_diabetes == 1) * 
                           df2_int$Y_qol) -
                      mean(wt * as.numeric(df2_int$A0_PM2.5==0 & 
                                             df2_int$M_diabetes == 1) * 
                             df2_int$Y_qol))
CDE_IPTW_m1_qol
# [1] -8.185866
```

### Stabilized IPTW for the CDE
If the controlled dired effect (CDE) is identifiable, $\Psi^{CDE}$ can be estimated using a stabilized IPTW estimator (modified Horvitz Thompson estimator):
\begin{equation}
\small
\hat{\mathbb{E}}(Y_{1,m}) - \hat{\mathbb{E}}(Y_{0,m}) =  \frac{ \sum_{i=1}^n \frac{\mathbb{I}(A_i=1 \cap M_i=m)}{\hat{g}(A_i=1 \mid L(0)_i) \times \hat{g}(M_i=m \mid L(0)_i,A_i,L(1)_i)} Y_i}{\sum_{i=1}^n \frac{\mathbb{I}(A_i=1 \cap M_i=m)}{\hat{g}(A_i=1 \mid L(0)_i) \hat{g}(M_i=m \mid L(0)_i,A_i,L(1)_i)}} - \frac{ \sum_{i=1}^n \frac{\mathbb{I}(A_i=0 \cap M_i=m)}{\hat{g}(A_i=0 \mid L(0)_i) \times \hat{g}(M_i=m \mid L(0)_i,A_i,L(1)_i)} Y_i}{ \sum_{i=1}^n \frac{\mathbb{I}(A_i=0 \cap M_i=m)}{\hat{g}(A_i=0 \mid L(0)_i) \times \hat{g}(M_i=m \mid L(0)_i,A_i,L(1)_i)}}
\end{equation}
The estimation algorithm is the same as for IPTW, but applying the stabilized estimator in step 4.

```{r Psi_CDE_siptw, echo=TRUE, eval = FALSE}
## 4. Applying the stabilized estimator
# point estimates of CDE, setting M=0:
CDE_sIPTW_m0_death <- (mean(wt * as.numeric(df2_int$A0_PM2.5 == 1 & 
                                              df2_int$M_diabetes == 0) * 
                              df2_int$Y_death) / 
                         mean(wt * as.numeric(df2_int$A0_PM2.5 == 1 & 
                                                df2_int$M_diabetes == 0))) -
                         (mean(wt * as.numeric(df2_int$A0_PM2.5 == 0 & 
                                              df2_int$M_diabetes == 0) * 
                              df2_int$Y_death) / 
                         mean(wt * as.numeric(df2_int$A0_PM2.5 == 0 & 
                                                df2_int$M_diabetes == 0)))
CDE_sIPTW_m0_death
# [1] 0.0601292

CDE_sIPTW_m0_qol <- (mean(wt * as.numeric(df2_int$A0_PM2.5 == 1 & 
                                              df2_int$M_diabetes == 0) * 
                              df2_int$Y_qol) / 
                         mean(wt * as.numeric(df2_int$A0_PM2.5 == 1 & 
                                                df2_int$M_diabetes == 0))) -
                         (mean(wt * as.numeric(df2_int$A0_PM2.5 == 0 & 
                                              df2_int$M_diabetes == 0) * 
                              df2_int$Y_qol) / 
                         mean(wt * as.numeric(df2_int$A0_PM2.5 == 0 & 
                                                df2_int$M_diabetes == 0)))
CDE_sIPTW_m0_qol
# [1] -4.966328

# point estimates of CDE, setting M=1:
CDE_sIPTW_m1_death <- (mean(wt * as.numeric(df2_int$A0_PM2.5 == 1 & 
                                              df2_int$M_diabetes == 1) * 
                              df2_int$Y_death) / 
                         mean(wt * as.numeric(df2_int$A0_PM2.5 == 1 & 
                                                df2_int$M_diabetes == 1))) -
                         (mean(wt * as.numeric(df2_int$A0_PM2.5 == 0 & 
                                              df2_int$M_diabetes == 1) * 
                              df2_int$Y_death) / 
                         mean(wt * as.numeric(df2_int$A0_PM2.5 == 0 & 
                                                df2_int$M_diabetes == 1)))
CDE_sIPTW_m1_death
# [1] 0.09030186

CDE_sIPTW_m1_qol <- (mean(wt * as.numeric(df2_int$A0_PM2.5 == 1 & 
                                              df2_int$M_diabetes == 1) * 
                              df2_int$Y_qol) / 
                         mean(wt * as.numeric(df2_int$A0_PM2.5 == 1 & 
                                                df2_int$M_diabetes == 1))) -
                         (mean(wt * as.numeric(df2_int$A0_PM2.5 == 0 & 
                                              df2_int$M_diabetes == 1) * 
                              df2_int$Y_qol) / 
                         mean(wt * as.numeric(df2_int$A0_PM2.5 == 0 & 
                                                df2_int$M_diabetes == 1)))
CDE_sIPTW_m1_qol
# [1] -10.03045
```


## Estimation of "Conditional" Randomized/Interventional Natural Direct (CRDE) and Indirect Effects (CRIE)

### IPTW for the CRDE and CRIE
Zheng and van der Laan [@zheng2017longitudinal] described "Conditional" Randomized Direct and Indirect Effects, using random draws from the counterfactual distribution of the mediator, conditional on both $L(0)$ and $L(1)$. The Average Total Effect (ATE) can be decomposed into the sum of :

  - a Conditional Randomized Natural Direct Effect (CRDE):
    $\text{CRDE}=\mathbb{E}(Y_{1,\Gamma_{0} \mid L(0),L(1)})-\mathbb{E}(Y_{0,\Gamma_{0} \mid L(0), L(1)})$,
  - and Conditional Randomized Natural Indirect Effect (CRIE)
    $\text{CRIE}=\mathbb{E}(Y_{1,\Gamma_{1} \mid L(0), L(1)})-\mathbb{E}(Y_{1,\Gamma_{0} \mid L(0), L(1)})$.

Under the identifiability conditions, the quantity of $\mathbb{E}(Y_{a,\Gamma_{a^\prime} \mid L(0), L(1)})$ can be estimated by IPTW:
\begin{equation*}
  \Psi^{a,a^\prime}_\text{IPTW} = \mathbb{E}\left[Y_{a,\Gamma_{a^\prime} \mid L(0), L(1)} \right] = \frac{1}{n} \sum_{i=1}^n D_\text{IPTW}^{a,a^\prime} \text{ where } D_\text{IPTW}^{a,a^\prime} = Y \frac{I(A=a)}{p_A{a\mid L(0)}} \frac{p_M(M \mid A=a^\prime,L(1),L(0))}{p_M(M \mid A=a,L(1),L(0))}
\end{equation*}

and the variance of the estimator $\Psi^{a,a^\prime}_\text{IPTW}$ is $\text{var}\left(\Psi^{a,a^\prime}_\text{IPTW}\right) = \frac{\text{var}\left(D_\text{IPTW}^{a,a^\prime}\right)}{n}$, which can be used to estimate 90\% confidence intervals.

```{r IPTW_CRDE_CRIE, echo=TRUE, eval = FALSE}
rm(list=ls())
df2_int <- read.csv(file = "./data/df2_int.csv")

## In order to estimate Psi^{a,a'}
## 1. Estimate gA and gM
gA <- glm(A0_PM2.5 ~ L0_male + L0_soc_env,
          family = "binomial", data = df2_int)
gM <- glm(M_diabetes ~ L0_male + L0_soc_env + A0_PM2.5 + L1,
          family = "binomial", data = df2_int)

## 2. Predict each individual's probability of being exposed to her own exposure
# predict the probabilities P(A0_PM2.5=1|L(0)) & P(A0_PM2.5=0|L(0))
pred.gA1 <- predict(gA, type = "response")
pred.gA0 <- 1 - pred.gA1
# the predicted probability of the observed treatment A_i=a is :
gAobs <- rep(NA, nrow(df2_int))
gAobs[df2_int$A0_PM2.5 == 1] <- pred.gA1[df2_int$A0_PM2.5 == 1]
gAobs[df2_int$A0_PM2.5 == 0] <- pred.gA0[df2_int$A0_PM2.5 == 0]

# predict the probabilities
# P(M=1|L(0),A=1,L(1)) & P(M=0|L(0),A=1,L(1)) setting A = 1
# and P(M=1|L(0),A=0,L(1)) & P(M=0|L(0),A=0,L(1)) setting A = 0
data.Ais0 <- data.Ais1 <- df2_int
data.Ais0$A0_PM2.5 <- 0
data.Ais1$A0_PM2.5 <- 1

pred.gM1.Ais1 <- predict(gM, newdata = data.Ais1, type = "response")
pred.gM0.Ais1 <- 1 - pred.gM1.Ais1
# the predicted probability of the observed treatment M_i=m is :
gMobs.Ais1 <- rep(NA, nrow(df2_int))
gMobs.Ais1[df2_int$M_diabetes == 1] <- pred.gM1.Ais1[df2_int$M_diabetes == 1]
gMobs.Ais1[df2_int$M_diabetes == 0] <- pred.gM0.Ais1[df2_int$M_diabetes == 0]

pred.gM1.Ais0 <- predict(gM, newdata = data.Ais0, type = "response")
pred.gM0.Ais0 <- 1 - pred.gM1.Ais0
# the predicted probability of the observed treatment M_i=m is :
gMobs.Ais0 <- rep(NA, nrow(df2_int))
gMobs.Ais0[df2_int$M_diabetes == 1] <- pred.gM1.Ais0[df2_int$M_diabetes == 1]
gMobs.Ais0[df2_int$M_diabetes == 0] <- pred.gM0.Ais0[df2_int$M_diabetes == 0]

## 3. Calculate D^{a,a'} - influence curve of the IPTW estimator
D.death.11 <- (df2_int$Y_death * (I(df2_int$A0_PM2.5 == 1) / gAobs) *
                 (gMobs.Ais1 / gMobs.Ais1))
D.death.10 <- (df2_int$Y_death * (I(df2_int$A0_PM2.5 == 1) / gAobs) *
                 (gMobs.Ais0 / gMobs.Ais1))
D.death.00 <- (df2_int$Y_death * (I(df2_int$A0_PM2.5 == 0) / gAobs) *
                 (gMobs.Ais0 / gMobs.Ais0))

D.qol.11 <- (df2_int$Y_qol * (I(df2_int$A0_PM2.5 == 1) / gAobs) *
               (gMobs.Ais1 / gMobs.Ais1))
D.qol.10 <- (df2_int$Y_qol * (I(df2_int$A0_PM2.5 == 1) / gAobs) *
               (gMobs.Ais0 / gMobs.Ais1))
D.qol.00 <- (df2_int$Y_qol * (I(df2_int$A0_PM2.5 == 0) / gAobs) *
               (gMobs.Ais0 / gMobs.Ais0))

## 4. Calculate CRDE and CRIE
## For death
CRDE.death <- mean(D.death.10) - mean(D.death.00)
CRDE.death
# [1] 0.07405068
CRIE.death <- mean(D.death.11) - mean(D.death.10)
CRIE.death
# [1] 0.00819879

## For quality of life
CRDE.qol <- mean(D.qol.10) - mean(D.qol.00)
CRDE.qol
# [1] -7.563116
CRIE.qol <- mean(D.qol.11) - mean(D.qol.10)
CRIE.qol
# [1] -0.8736813

## 5. Calculate 95% CI based on the influence curve D^{a,a'}
# the variance of the estimator Psi^{a,a'} = mean(D^{a,a'}) is var(D^{a,a'}) / n)
se.CRDE.death <- sqrt(var(D.death.10 - D.death.00) / nrow(df2_int))
c(CRDE.death - qnorm(0.975) * se.CRDE.death,
  CRDE.death + qnorm(0.975) * se.CRDE.death)
# [1] 0.04141535 0.10668602
se.CRIE.death <- sqrt(var(D.death.11 - D.death.10) / nrow(df2_int))
c(CRIE.death - qnorm(0.975) * se.CRIE.death,
  CRIE.death + qnorm(0.975) * se.CRIE.death)
# [1] 0.003380238 0.013017342

se.CRDE.qol <- sqrt(var(D.qol.10 - D.qol.00) / nrow(df2_int))
c(CRDE.qol - qnorm(0.975) * se.CRDE.qol,
  CRDE.qol + qnorm(0.975) * se.CRDE.qol)
# [1] -11.748095  -3.378137
se.CRIE.qol <- sqrt(var(D.qol.11 - D.qol.10) / nrow(df2_int))
c(CRIE.qol - qnorm(0.975) * se.CRIE.qol,
  CRIE.qol + qnorm(0.975) * se.CRIE.qol)
# [1] -1.4111404 -0.3362223
```
Results are close to the estimations obtained by g-computation. IPTW estimations are know to be more sensitive to positivity issues, with larger confidence intervals (they can be too conservative with more than 95% coverage). 

  - the conditional "randomized" Natural Indirect effect is $\text{CRIE} \approx +0.8\%, \quad 95\%CI=[0.3\%,1.3\%]$ on death and $\text{CRIE} \approx -0.9, \quad 95\%CI=[-1.4,-0.3]$ on quality of life. This indirect effect corresponds to the specific path $A \rightarrow M \rightarrow Y$ (and can also contain the mediated interactive effect of due to the $A \ast M$ interaction on $Y$).
  - the conditional "randomized" Natural Direct effect is $\text{CRDE} \approx +7.4\, \quad 95\%CI=[4.1\%,10.7\%]$ on death and $\text{CRDE} \approx -7.6, \quad 95\%CI=[-11.7,-3.4]$ on quality of life. This direct effect corresponds to the combination of the paths $A \rightarrow Y$, $A \rightarrow L(1) \rightarrow Y$ and $A \rightarrow L(1) \rightarrow M \rightarrow Y$. 

95% confidence intervals can be calculated by bootstrap.

