# Marginal structural models {#msm_chapter}

Marginal structural models (MSM) are parametric models that are used to summarize the relationship between the counterfactual outcome ($Y_a$ or $Y_{am}$ for example) and the exposure(s) $A$ and mediators $M$. It also possible to summarize the relationship according to a subset of the baseline confounders if it is relevant for the scientific question.

To illustrate the application of MSMs, we will first use the data simulated from the Causal model 1 (with an $A \ast M$ interaction effect on the outcome) where the exposure $A$ doesn't affect the counfonder $L(1)$ between the mediator and the outcome.

## MSM for the Average Total Effect (ATE) {#msm_ATE_paragraph}

### Expressing the ATE using coefficients of an MSM
For a continuous or binary outcome, we can use the following MSM to summarize the relationship between the counterfactual outcome ($Y_a$) and the exposure(s) $A$:
\begin{equation} 
  \mathbb{E}(Y_a) = \alpha_0 + \alpha_A a 
  (\#eq:MSMATEmarg)
\end{equation}

The Average Total Effect $\text{ATE} = \mathbb{E}(Y_{A=1}) - \mathbb{E}(Y_{A=0})$ can then be expressed using the coefficients of this MSM \@ref(eq:MSMATEmarg):
\begin{equation*} 
  \text{ATE} := \left(\alpha_0 + \alpha_A \times 1 \right) - \left(\alpha_0 + \alpha_A \times 0 \right) = \alpha_A 
\end{equation*}

In this example, the coefficient $\alpha_A$ corresponds to the $\text{ATE}$. 

Such a model is not very useful for a binary exposure. It would be much more useful for higher-dimensional exposures, for example with a continuous exposure, where the relationship between all the possible continuous values of the exposure $A=a$ and the corresponding outcomes $Y_a$ is summarized (and arbitrarily simplified) by a single line and the slope coefficient $\alpha_A$. 


It is also possible to define MSMs adjusted for a subset $V$ of the baseline confounders ($V \subset L(0)$). Such MSMs can be useful to estimate conditional effects. For example it is possible to define an average "conditional" total effect $\text{ATE}|L(0)$ (instead of the marginal ATE defined above), projecting the counterfactual outcomes on a parametric model such as the following: 

\begin{equation} 
  \mathbb{E}(Y_a \mid L(0)) = \alpha_0 + \alpha_A a + \alpha_\text{male} L_\text{male}(0) + \alpha_\text{low.educ.} L_\text{low.educ.}(0)
  (\#eq:MSMATEcond)
\end{equation}

so that $\text{ATE} \mid L(0) = \mathbb{E}(Y_{A=1} \mid L(0)) - \mathbb{E}(Y_{A=0} \mid L(0)) = \alpha_A$ using the coefficient from the MSM \@ref(eq:MSMATEcond).


MSMs are also very useful to study interactions, or effect modification of the exposure $A$ by a baseline confounder. For example, in order to study the average total effect according to sex, we can use the following MSM:

\begin{equation} 
  \mathbb{E}(Y_a \mid L_\text{male}(0)) = \alpha_0 + \alpha_A a + \alpha_\text{male} L_\text{male}(0) + \alpha_{A \ast L_\text{male}} \left(a \times L_\text{male}(0)\right)
  (\#eq:MSMATEsex)
\end{equation}

and express the average total effect in each strata of sex using the coefficients of the MSM \@ref(eq:MSMATEsex):

\begin{align*}
 \{\text{ATE} \mid L_\text{male}(0) = 0\} &:= \mathbb{E}(Y_1 \mid L_\text{male}(0) = 0) - \mathbb{E}(Y_0 \mid L_\text{male}(0) = 0) = \alpha_A \\
 \{\text{ATE} \mid L_\text{male}(0) = 1\} &:= \mathbb{E}(Y_1 \mid L_\text{male}(0) = 1) - \mathbb{E}(Y_0 \mid L_\text{male}(0) = 1) = \alpha_A + \alpha_{A \ast L_\text{male}}
\end{align*}

Because MSMs are models of unobserved counterfactual outcomes, estimators of the MSM coefficients are necessary. We will describe two possible approaches : estimation by IPTW or by G-computation.

In both approaches, 95\% confidence intervals can be computed by bootstrap.

### Estimation of the MSM coefficients by IPTW
MSM coefficients can be easily estimated using an Inverse Probability of Treatment (IPTW) approach based on weighted regressions. 

For example, in order to fit the MSM \@ref(eq:MSMATEsex) described above, we can use a linear regression of the (observed) outcome $Y$ on the exposure and sex, weigthed by individual weights $w_i$ or $sw_i$:
\begin{equation} 
  \mathbb{E}\left[Y \mid L_\text{male}(0)\right] = \alpha_0 + \alpha_A a + \alpha_\text{male} L_\text{male}(0) + \alpha_{A \ast L_\text{male}} \left(a \times L_\text{male}(0)\right)
\end{equation}

where $w_i=\frac{1}{P(A=a_i \mid L(0)=l(0)_i)}$ or $sw_i=\frac{P(A=a_i \mid L_\text{male}(0))}{P(A=a_i \mid L(0)=l(0)_i)}$. 

As in chapter \@ref(ChapIptw), the "no-unmeasured confounding" assumption is addressed by the application of weights $w_i$ or $sw_i$, which balance confounders $L(0)$ relative to the exposure $A$.

```{r Psi_ATE_bysex_MSM, echo=TRUE, eval = FALSE}
## 1. Denominator of the weight
# 1a. Estimate g(A=a_i|L(0)) (denominator of the weight)
g.A.L <- glm(A0_ace ~ L0_male + L0_parent_low_educ_lv,
           family = "binomial", data = df1_int)

# 1b. Predict each individual's probability of being exposed to her own exposure
# predict the probabilities P(A0_ace=1 & P(A0_ace=0)
pred.g1.L <- predict(g.A.L, type="response")
pred.g0.L <- 1 - pred.g1.L
# the predicted probability of the observed treatment P(A = a_i | L(0)) is :
gAi.L <- rep(NA, nrow(df1_int))
gAi.L[df1_int$A0_ace==1] <- pred.g1.L[df1_int$A0_ace==1]
gAi.L[df1_int$A0_ace==0] <- pred.g0.L[df1_int$A0_ace==0]

## 2. Numerator of the weight
# The numerator of the weight can be 1 for simple weights,
# or g(A=a_i|V) to obtain stabilized weights which put less weight to individuals
# with less observation. Stabilized weights enable a weaker positivity assumption.

# 2a. Estimate g(A=a_i | sex) (numerator of the stabilized weight)
g.A.sex <- glm(A0_ace ~ L0_male,
           family = "binomial", data = df1_int)

# 2b. Predict each individual's probability of being exposed to her own exposure
# predict the probabilities P(A0_ace=1 | sex) & P(A0_ace=0 | sex)
pred.g1.sex <- predict(g.A.sex, type="response")
pred.g0.sex <- 1 - pred.g1.sex
# the predicted probability of the observed treatment P(A = a_i | sex) is :
gAi.sex <- rep(NA, nrow(df1_int))
gAi.sex[df1_int$A0_ace==1] <- pred.g1.sex[df1_int$A0_ace==1]
gAi.sex[df1_int$A0_ace==0] <- pred.g0.sex[df1_int$A0_ace==0]

## 3. Define individual weights:
# We can use simple weights w = 1 / g(A=a_i | L(0))
w <- 1 / gAi.L
# Or alternatively, we can use stabilized weights : 
# sw = g(A=a_i | sex) / g(A=a_i | L(0))
sw <- gAi.sex / gAi.L

# we can see that stabilized weights have less extreme values
par(mfcol = c(1,2))
boxplot(w ~ df1_int$A0_ace)
boxplot(sw ~ df1_int$A0_ace)

## 4. Estimate coefficients of the MSM using a weighted regression E(Y | A, sex)
# a GLM with gaussian family can be applied to estimate risk differences
# (for relative risk or rate ratios, we can apply a Poisson family; 
#  for OR, we can apply a binomial family)
msm1 <- glm(Y_death ~ A0_ace + L0_male + A0_ace*L0_male,
           weights = w, # applying the simple weight
           family = "gaussian",
           data = df1_int)
coef(msm1)
# (Intercept)         A0_ace        L0_male A0_ace:L0_male
# 0.17573472     0.03589627     0.04598911     0.04136896

msm2 <- glm(Y_death ~ A0_ace + L0_male + A0_ace*L0_male,
            weights = sw, # applying the stabilized weight
            family = "gaussian",
            data = df1_int)
coef(msm2)
# (Intercept)         A0_ace        L0_male A0_ace:L0_male
# 0.17573472     0.03589627     0.04598911     0.04136896

## 5. Estimate the ATE stratified by sex
# According to the MSM1 (with simple weights)
ATE.msm1.male0 <- coef(msm1)["A0_ace"]
# 0.03589627
ATE.msm1.male1 <- coef(msm1)["A0_ace"] + coef(msm1)["A0_ace:L0_male"]
# 0.07726522

# According to the MSM2 (with stabilized weights)
ATE.msm2.male0 <- coef(msm2)["A0_ace"]
# 0.03589627
ATE.msm2.male1 <- coef(msm2)["A0_ace"] + coef(msm2)["A0_ace:L0_male"]
# 0.07726522
# The results are the same because there is no violation of the positivity assumption
# In case of positivity violation, stabilized weights would give more accurate estimates
```

The ATE estimates of death probability using an MSM estimated by IPTW are respectively +3.6\% in women and +7.7\% in men.

95\% confidence intervals can be calculated by bootstrap.

Note: Using the true data generating model used to simulate the illustrative datasets, the "true" value of the ATE stratified by sex can be calculated: 

  - the "true" $(ATE \mid L_\text{male}(0) = 0) = 0.0688$ in women,

  - the "true" $(ATE \mid L_\text{male}(0) = 1) = 0.0703$ in men.

### Estimation of the MSM coefficients by G-computation (imputation)
We can also use a G-computation (sometimes described as an imputation) approach to estimate the coefficients of an MSM.

The following steps can be applied:

1. Fit a (logistic or a linear) regression to estimate $\overline{Q} = \mathbb{E}(Y \mid A, L(0))$

2. Use this estimate to predict an outcome for each subject under the counterfactual scenarios $\hat{\overline{Q}}(A=0)_i$ and $\hat{\overline{Q}}(A=1)_i$, by evaluating the regression fit $\overline{Q}$ at $A=0$ and $A=1$ respectively

3. Duplicate the initial dataset in a single long dataset in which:

  - the first half of the long dataset corresponds to the first counterfactual scenario with $A=0$ for all individuals and an additional column for the predicted counterfactual outcome $\hat{\overline{Q}}(A=0)$ ; 
  - the second half of the long dataset corresponds to the second counterfactual scenario with $A=1$ for all individuals and $\hat{\overline{Q}}(A=1)$ for the predicted counterfactual column.

4. Fit the MSM $\mathbb{E}\left[Y_a \mid L_\text{male}(0)\right]$ using the long dataset.


```{r Psi_ATE_bysex_MSM_gcomp, echo=TRUE, eval = FALSE}
## 1. Estimate Qbar
Q.tot.death <- glm(Y_death ~ A0_ace + L0_male + L0_parent_low_educ_lv,
                   family = "gaussian", data = df1_int)
# The final result would be sligthly different if we apply a binomial family.

## 2. Predict an outcome for each subject, in 2 counterfactual scenarios 
##    setting A=0 and A=1
# prepare data sets used to predict the outcome under the counterfactual
# scenarios setting A=0 and A=1
data.A1 <- data.A0 <- df1_int
data.A1$A0_ace <- 1
data.A0$A0_ace <- 0

# predict values under the same name in the corresponding counterfactual dataset
data.A1$Ya.death.pred <- predict(Q.tot.death, newdata = data.A1, type = "response")
data.A0$Ya.death.pred <- predict(Q.tot.death, newdata = data.A0, type = "response")

## 3. Append both counterfactual datasets in a single long dataset
# (the number of row is twice the initial number of row because there are 2 
# counterfactual scenarios)
data.2scenarios <- rbind(data.A0, data.A1)

## 4. fit the MSM: E(Y_a|sex)
# a GLM with gaussian family can be applied to estimate risk differences
MSM.ATE.gcomp <- glm(Ya.death.pred ~ A0_ace + L0_male + A0_ace:L0_male,
                     family = "gaussian",
                     data = data.2scenarios)
coef(MSM.ATE.gcomp)
#  (Intercept)         A0_ace        L0_male A0_ace:L0_male
# 1.743994e-01   7.720726e-02   4.874750e-02   4.802530e-16

## 5. Estimate the ATE stratified by sex
# According to MSM.ATE.gcomp
ATE.MSM.gcomp.male0 <- coef(MSM.ATE.gcomp)["A0_ace"]
# 0.07720726
ATE.MSM.gcomp.male1 <- (coef(MSM.ATE.gcomp)["A0_ace"] +
                          coef(MSM.ATE.gcomp)["A0_ace:L0_male"])
# 0.07720726
# The results are the same in both strata, because in the first Qbar model,
# we did not include any (A * sex) interaction term

# Applying a binomial family for the first Qbar model would result in two
# different values of the ATE stratified by sex.
# => 0.06880798 in the L0_male = 0 strata
# => 0.08053575 in the L0_male = 1 strata

# Comments: For the estimation of the first Qbar model, applying a gaussian family 
# (additive model) with no interaction terms implies the presence of some interaction 
# terms in a multiplicative model (such as a glm with a binomial family).
# On the contrary, applying a binomial family (multiplicative model) with no
# interaction terms implies some interaction terms in an additive model (such as
# a glm with a gaussian family)
```

## MSM for Controlled Direct Effects {#msm_CDE_paragraph}
### Expressing the CDE using coefficients of an MSM

The controlled direct effect is defined as $\text{CDE}_m= \mathbb{E}(Y_{am}) - \mathbb{E}(Y_{a^*m})$.

Using the following MSM 
\begin{equation} 
  \mathbb{E}(Y_{am}) = \alpha_0 + \alpha_A a + \alpha_M m + \alpha_{A \ast M} a \times m
  (\#eq:MSMCDE)
\end{equation}
the controlled direct effect (keeping the mediator constant to the value $M=m$) can be expressed using the coefficients of the MSM \@ref(eq:MSMCDE):
\begin{align*} 
  \text{CDE}_m &= (\alpha_0 + \alpha_A a + \alpha_M m + \alpha_{A \ast M} a \times m) - (\alpha_0 + \alpha_A a^* + \alpha_M m + \alpha_{A \ast M} a^* \times m) \\
  \text{CDE}_m &= \alpha_A(a - a^* ) + \alpha_{A \ast M} \times (a - a^*) \times m
\end{align*}
For a binary exposure $A$, we have $\text{CDE}_m=\alpha_A + \alpha_{A \ast M} \times m$.

### Estimation of the MSM coefficients by IPTW
MSM coefficients can be easily estimated using an Inverse Probability of Treatment (IPTW) approach based on weighted regressions. 

In order to fit the MSM \@ref(eq:MSMCDE), we can use a linear regression of the (observed) outcome $Y$ on the exposure and mediator, weighted by individual stabilized weights $sw_i$ ([@vanderweele2009]):
\begin{equation} 
  \mathbb{E}\left(Y \mid A,M\right) = \alpha_0 + \alpha_A a + \alpha_M m + \alpha_{A \ast M} a \times m
\end{equation}

where $sw_i$ is the product of two weights $sw_i = sw_{A,i} \times sw_{M,i}$, 

$sw_{A,i}=\frac{P(A=a_i)}{P(A=a_i \mid L(0)=l(0)_i)}$ and $sw_{M,i}=\frac{P(M=m_i \mid A=a_i)}{P(M = m_i \mid A=a_i,L(0)=l(0)_i), L(1)=l(1)_i}$. 

The "no-unmeasured confounding" assumption is addressed by the application of weights  $sw_i$, which balance confounders $L(0)$ relative to the exposure-outcome $A-Y$ relationship, and balance the set of confounders $\{L(0),A,L(1)\}$ relative to the mediator-outcome $M-Y$ relationship. 

Importantly, this approach for the estimation of the controlled direct effect $\text{CDE}_m$ by IPTW is also valid if the exposure $A$ affects the intermediate confounder $L(1)$ (as with the Causal model 2).

```{r Psi_CDE_MSM_IPTW, echo=TRUE, eval = FALSE}
### MSM of CDE, estimated by IPTW ----------------------------------------------
## 1. Stabilized weight for the exposure sw_{A,i}
# 1a. Estimate g(A=a_i|L(0)) (denominator of the weight)
g.A.L <- glm(A0_ace ~ L0_male + L0_parent_low_educ_lv,
             family = "binomial", data = df1_int)
# 1b. Predict each individual's probability of being exposed to her own exposure
# the predicted probability of the observed treatment g(A = a_i | L(0)) is :
gAi.L <- rep(NA, nrow(df1_int))
gAi.L[df1_int$A0_ace==1] <- predict(g.A.L, type="response")[df1_int$A0_ace==1]
gAi.L[df1_int$A0_ace==0] <- (1 - predict(g.A.L, type="response"))[df1_int$A0_ace==0]

# 1c. Estimate g(A=a_i) (numerator of the weight)
g.A <- glm(A0_ace ~ 1, family = "binomial", data = df1_int)
# 1d. Predict each individual's probability of being exposed to her own exposure
# the predicted probability of the observed treatment g(A = a_i) is :
gAi <- rep(NA, nrow(df1_int))
gAi[df1_int$A0_ace==1] <- predict(g.A, type="response")[df1_int$A0_ace==1]
gAi[df1_int$A0_ace==0] <- (1 - predict(g.A, type="response"))[df1_int$A0_ace==0]

# 1e. Calculate the weight for the exposure A: sw_{A,i}
sw_Ai <- gAi / gAi.L

## 2. Stabilized weight for the mediator sw_{M,i}
# 2a. Estimate g(M=m_i|L(0),A,L(1)) (denominator of the weight)
g.M.L <- glm(M_smoking ~ L0_male + L0_parent_low_educ_lv + A0_ace + L1,
             family = "binomial", data = df1_int)
# 2b. Predict each individual's probability of being exposed to her own exposure
# the predicted probability of the observed treatment g(A = a_i | L(0)) is :
gMi.L <- rep(NA, nrow(df1_int))
gMi.L[df1_int$M_smoking==1] <- predict(g.M.L, type="response")[df1_int$M_smoking==1]
gMi.L[df1_int$M_smoking==0] <- (1 - predict(g.M.L, type="response"))[df1_int$M_smoking==0]

# 2c. Estimate g(M=m_i|A) (numerator of the weight)
g.M.A <- glm(M_smoking ~ A0_ace, family = "binomial", data = df1_int)
# 2d. Predict each individual's probability of being exposed to her own exposure
# the predicted probability of the observed treatment g(M = m_i|A) is :
gMi.A <- rep(NA, nrow(df1_int))
gMi.A[df1_int$M_smoking==1] <- predict(g.M.A, type="response")[df1_int$M_smoking==1]
gMi.A[df1_int$M_smoking==0] <- (1 - predict(g.M.A, type="response"))[df1_int$M_smoking==0]
# 2e. Calculate the weight for the mediator M: sw_{M,i}
sw_Mi <- gMi.A / gMi.L

## 3. Define the individual stabilized weight for the CDE_m
sw_cde <- sw_Ai * sw_Mi

## 4. Estimate coefficients of the MSM using a weighted regression E(Y | A, sex)
# a GLM with gaussian family can be applied to estimate risk differences
msm_cde <- glm(Y_death ~ A0_ace + M_smoking + A0_ace*M_smoking,
               weights = sw_cde,
               family = "gaussian",
               data = df1_int)
coef(msm_cde)
# (Intercept)           A0_ace        M_smoking A0_ace:M_smoking 
#  0.17891689       0.06798282       0.06729724      -0.00495314

## 5. Estimate CDE for m=0 and for m=1 using the MSM's coefficients
CDE_mis0 <- coef(msm_cde)["A0_ace"]
# 0.06798282
CDE_mis1 <- coef(msm_cde)["A0_ace"] + coef(msm_cde)["A0_ace:M_smoking"]
# 0.06302968
```

In this example, our estimates of the controlled direct effects are $CDE_{M=0} = 6.8\%$ and $CDE_{M=1} = 6.3\%$. Confidence intervals can be calculated by bootstrap.


### Estimation of the MSM coefficients by G-computation
As for the ATE, we can use G-computation to estimate the coefficients of the MSM to estimate Controlled Direct Effects.

Note that the algorithm described below is correct only if the exposure $A$ doesn't affect the intermediate confounders $L(1)$ of the $M \rightarrow Y$ relationship (such as the data simulated from the Causal model 1 in our examples). In that case, the following steps can be applied:

1. Fit a (logistic or a linear) regression to estimate $\overline{Q} = \mathbb{E}(Y \mid A,M, L(0),L(1))$

2. Use this estimate to predict an outcome for each subject under the counterfactual scenarios $\hat{\overline{Q}}(A=0,M=m,L(0),L(1))_i$ and $\hat{\overline{Q}}(A=1,M=m,L(0),L(1))_i$, by evaluating the regression fit $\overline{Q}$ at $(A=0,M=m)$ and $(A=1,M=m)$ respectively. If we want to set the level of the mediator to $M=0$ and $M=1$, this would give 4 counterfactual scenarios $\text{do}(A=0,M=0)$, $\text{do}(A=1,M=0)$, $\text{do}(A=0,M=1)$ and $\text{do}(A=1,M=1)$. 

3. Duplicate the initial dataset for each scenario in a single long dataset in which:

  - the 1st part of the long dataset corresponds to the first counterfactual scenario with $(A=0,M=0)$ for all individuals and an additional column for the predicted counterfactual outcome $\mathbb{E}(Y_{A=0,M=0}\mid L(0),L(1)) = \hat{\overline{Q}}(A=0,M=0,L(0),L(1))$ ; 
  - the 2d part of the long dataset corresponds to the second counterfactual scenario with $(A=1,M=0)$ for all individuals and an additional column for the predicted counterfactual outcome $\mathbb{E}(Y_{A=1,M=0}\mid L(0),L(1)) = \hat{\overline{Q}}(A=1,M=0,L(0),L(1))$ ; 
  - the 3d part of the long dataset corresponds to the second counterfactual scenario with $(A=0,M=1)$ for all individuals and an additional column for the predicted counterfactual outcome $\mathbb{E}(Y_{A=0,M=1}\mid L(0),L(1)) = \hat{\overline{Q}}(A=0,M=1,L(0),L(1))$ ; 
  - the 4th part of the long dataset corresponds to the second counterfactual scenario with $(A=1,M=1)$ for all individuals and an additional column for the predicted counterfactual outcome $\mathbb{E}(Y_{A=1,M=1}\mid L(0),L(1)) = \hat{\overline{Q}}(A=1,M=1,L(0),L(1))$ ; 

4. Fit the MSM $\mathbb{E}(Y_{am}) = \alpha_0 + \alpha_A a + \alpha_M m + \alpha_{A \ast M} a \times m$ using the long dataset.

```{r Psi_CDE_MSM_gcomp, echo=TRUE, eval = FALSE}
### MSM of CDE, estimated by G-computation -------------------------------------
## 1. Estimate Qbar(A,M,L0,L1)
Q.cde.death <- glm(Y_death ~ A0_ace + M_smoking + A0_ace:M_smoking + L0_male + 
                     L0_parent_low_educ_lv + L1,
                   family = "gaussian", data = df1_int)
# The final result would be sligthly different if we applied a binomial family
# The Gaussian family corresponds to the true generating model in this example.

## 2. Predict an outcome for each subject, in each counterfactual scenario
# Prepare data sets that will be used to predict the outcome under the counterfactual
# 4 counterfactual scenarios setting (A=0,M=0), (A=1,M=0), (A=0,M=1) and (A=1,M=1)
data.A0M0 <- data.A1M0 <- data.A0M1 <- data.A1M1 <- df1_int
data.A0M0$A0_ace <- 0
data.A0M0$M_smoking <- 0

data.A1M0$A0_ace <- 1
data.A1M0$M_smoking <- 0

data.A0M1$A0_ace <- 0
data.A0M1$M_smoking <- 1

data.A1M1$A0_ace <- 1
data.A1M1$M_smoking <- 1

# predict values under the same name in the corresponding counterfactual dataset
data.A0M0$Yam.death.pred <- predict(Q.cde.death, newdata = data.A0M0, type = "response")
data.A1M0$Yam.death.pred <- predict(Q.cde.death, newdata = data.A1M0, type = "response")
data.A0M1$Yam.death.pred <- predict(Q.cde.death, newdata = data.A0M1, type = "response")
data.A1M1$Yam.death.pred <- predict(Q.cde.death, newdata = data.A1M1, type = "response")

## 3. Append the 4 counterfactual datasets in a single long dataset
# number of row is 4 times the initial value (we have 4 counterfactual scenarios)
data.4scenarios <- rbind(data.A0M0, data.A1M0,data.A0M1,data.A1M1)

## 4. fit the MSM: E(Y_am) = alpha_0 + alpha_A a + alpha_M m + alpha_AM a:m
MSM.CDE.gcomp <- glm(Yam.death.pred ~ A0_ace +  M_smoking + A0_ace:M_smoking,
                     family = "gaussian", # gaussian family for risk differences
                     data = data.4scenarios)
coef(MSM.CDE.gcomp)
# (Intercept)           A0_ace        M_smoking A0_ace:M_smoking
#  0.17968603       0.06000138       0.06757214       0.01918153

## 5. Estimate the CDE(M=m)
# CDE(M=0) = E(Y_{A=1,M=0}) - E(Y_{A=0,M=0})
CDE_mis0_gcomp <- coef(MSM.CDE.gcomp)["A0_ace"]
# 0.06000138

# CDE(M=1) = E(Y_{A=1,M=1}) - E(Y_{A=0,M=1})
CDE_mis1_gcomp <- (coef(MSM.CDE.gcomp)["A0_ace"] +
                     coef(MSM.CDE.gcomp)["A0_ace:M_smoking"])
# 0.07918291

# Note: Applying a binomial family for the first Qbar model would result in two
# sligthly different values of the CDE(M=m)
# => 0.05934409 in setting M=0
# => 0.07537874 in setting M=1
```

If the exposure $A$ affects the intermediate confounder $L(1)$, as in the Causal model 2, the steps 1) and 2) of the algorithm should follow the method described in paragraph \@ref(ChapGcomp-CDE-param) or \@ref(ChapGcomp-CDE-ICE). 

Here is an example applying G-computation by iterative conditional expectation:
```{r Psi_CDE_MSM_gcomp_ice, echo=TRUE, eval = FALSE}
### MSM of CDE, estimated by G-computation (by ICE) ----------------------------
## 1a) Regress the outcome on L0, A, L1 and M (and the A*M interaction if appropriate)
Y.death.model <- glm(Y_death ~ L0_male + L0_parent_low_educ_lv + A0_ace + L1 +
                       M_smoking + A0_ace:M_smoking,
                     family = "binomial", data = df2_int)

## 1b) Generate predicted values by evaluating the regression  
##     under the 4 counterfactual scenarios
data.A0M0 <- data.A1M0 <- data.A0M1 <- data.A1M1 <- df2_int
data.A0M0$A0_ace <- 0
data.A0M0$M_smoking <- 0

data.A1M0$A0_ace <- 1
data.A1M0$M_smoking <- 0

data.A0M1$A0_ace <- 0
data.A0M1$M_smoking <- 1

data.A1M1$A0_ace <- 1
data.A1M1$M_smoking <- 1

Q.Y.death.A0M0 <- predict(Y.death.model, newdata = data.A0M0, type = "response")
Q.Y.death.A1M0 <- predict(Y.death.model, newdata = data.A1M0, type = "response")
Q.Y.death.A0M1 <- predict(Y.death.model, newdata = data.A0M1, type = "response")
Q.Y.death.A1M1 <- predict(Y.death.model, newdata = data.A1M1, type = "response")

## 2a) Regress the predicted values conditional on the observed exposure A
##    and baseline confounders L(0)
L1.death.A0M0.model <- glm(Q.Y.death.A0M0 ~ L0_male + L0_parent_low_educ_lv + A0_ace,
                           family = "quasibinomial", data = df2_int)
L1.death.A1M0.model <- glm(Q.Y.death.A1M0 ~ L0_male + L0_parent_low_educ_lv + A0_ace,
                           family = "quasibinomial", data = df2_int)
L1.death.A0M1.model <- glm(Q.Y.death.A0M1 ~ L0_male + L0_parent_low_educ_lv + A0_ace,
                           family = "quasibinomial", data = df2_int)
L1.death.A1M1.model <- glm(Q.Y.death.A1M1 ~ L0_male + L0_parent_low_educ_lv + A0_ace,
                           family = "quasibinomial", data = df2_int)


## 2b) generate predicted values by evaluating the regression at exposure
##    of interest: {A=0,M=0}, {A=1,M=0}, {A=0,M=1}, {A=1,M=1}
data.A0M0$Yam.death.pred <- predict(L1.death.A0M0.model,
                                    newdata = data.A0M0, type = "response")
data.A1M0$Yam.death.pred <- predict(L1.death.A1M0.model,
                                    newdata = data.A1M0, type = "response")
data.A0M1$Yam.death.pred <- predict(L1.death.A0M1.model,
                                    newdata = data.A0M1, type = "response")
data.A1M1$Yam.death.pred <- predict(L1.death.A1M1.model,
                                    newdata = data.A1M1, type = "response")


## 3. Append the 4 counterfactual datasets in a single long dataset
# number of row is 4 times the initial value (we have 4 counterfactual scenarios)
data.4scenarios <- rbind(data.A0M0, data.A1M0,data.A0M1,data.A1M1)

## 4. fit the MSM: E(Y_am) = alpha_0 + alpha_A a + alpha_M m + alpha_AM a:m
MSM.CDE.gcomp <- glm(Yam.death.pred ~ A0_ace +  M_smoking + A0_ace:M_smoking,
                     family = "gaussian", # gaussian family for risk differences
                     data = data.4scenarios)
coef(MSM.CDE.gcomp)
# (Intercept)           A0_ace        M_smoking A0_ace:M_smoking
#  0.17974947       0.06342833       0.07366466       0.02469485

## 5. Estimate the CDE(M=m)
# CDE(M=0) = E(Y_{A=1,M=0}) - E(Y_{A=0,M=0})
CDE_mis0_gcomp_ice <- coef(MSM.CDE.gcomp)["A0_ace"]
# 0.06342833

# CDE(M=1) = E(Y_{A=1,M=1}) - E(Y_{A=0,M=1})
CDE_mis1_gcomp_ice <- (coef(MSM.CDE.gcomp)["A0_ace"] +
                         coef(MSM.CDE.gcomp)["A0_ace:M_smoking"])
# 0.08812318
```

The example above (MSM estimation using G-computation by ICE) corresponds to the algorithm applied by the `ltmle` package:

```{r Psi_CDE_MSM_ltmle_gcomp, echo=TRUE, eval = FALSE}
library(ltmle)
Qform <- c(L1="Q.kplus1 ~ L0_male + L0_parent_low_educ_lv + A0_ace",
           Y_death="Q.kplus1 ~ L0_male + L0_parent_low_educ_lv + L1 +
                    A0_ace * M_smoking")
gform <- c("A0_ace ~ L0_male + L0_parent_low_educ_lv",
           "M_smoking ~ L0_male + L0_parent_low_educ_lv + A0_ace + L1")
# in this example of g-computation, the propensity scores 'gform' will not be used

data_binary <- subset(df2_int, select = c(L0_male, L0_parent_low_educ_lv,
                                          A0_ace, L1,
                                          M_smoking, Y_death))
CDE_ltmle_M0 <- ltmle(data = data_binary,
                      Anodes = c("A0_ace", "M_smoking"),
                      Lnodes = c("L1"), # intermediate confounders +/- baseline
                      Ynodes = c("Y_death"),
                      survivalOutcome = FALSE, # TRUE for time-to-event outcomes Y
                      Qform = Qform,
                      gform = gform,
                      abar = list(c(1,0), # counterfactual intervention do(A=1,M=0)
                                  c(0,0)), # counterfactual intervention do(A=0,M=0)
                      SL.library = NULL, # call glm() instead of SuperLearner
                      estimate.time = FALSE, # estimate computation time?
                      gcomp = TRUE, # to apply g-computation
                      variance.method = "ic")
# CDE with M=0
summary(CDE_ltmle_M0)$effect.measures$ATE$estimate
#   Parameter Estimate:  0.06342833

CDE_ltmle_M1 <- ltmle(data = data_binary,
                      Anodes = c("A0_ace", "M_smoking"),
                      Lnodes = c("L1"), # intermediate confounders +/- baseline
                      Ynodes = c("Y_death"),
                      survivalOutcome = FALSE, # TRUE for time-to-event outcomes Y
                      Qform = Qform,
                      gform = gform,
                      abar = list(c(1,1), # counterfactual intervention do(A=1,M=0)
                                  c(0,1)), # counterfactual intervention do(A=0,M=0)
                      SL.library = NULL, # call glm() instead of SuperLearner
                      estimate.time = FALSE, # estimate computation time?
                      gcomp = TRUE, # to apply g-computation
                      variance.method = "ic")
# CDE with M=1
summary(CDE_ltmle_M1)$effect.measures$ATE$estimate
#   Parameter Estimate:  0.08812318
```


## MSM for Natural Direct and Indirect Effects {#msm_NDE_NIE_paragraph}
### Expressing the NDE and NIE using coefficients of 2 MSMs
The (Pure) Natural Direct Effect is defined by $\text{PNDE} = \mathbb{E}(Y_{a,M_{a^*}}) - \mathbb{E}(Y_{a^*,M_{a^*}})$ and the (Total) Natural Indirect Effect is defined by $\text{TNIE} = \mathbb{E}(Y_{a,M_a}) - \mathbb{E}(Y_{a,M_{a^*}})$.

VanderWeele suggested using 2 MSMs conditional on baseline confounders $L(0)$ in order to estimate natural direct and indirect effects [@vanderweele2009]: 

  1) a model of the counterfactual values of the outcome $\mathbb{E}(Y_{a,m} \mid l(0))=h^{-1}(a,m,l(0))$, where $h$ is a link function. For example:
  \begin{equation}
    \mathbb{E}(Y_{a,m} \mid l(0)) = \alpha_0 + \alpha_A a + \alpha_M m + \alpha_{AM} a \times m + \alpha_{L(0)} l(0) (\#eq:MSM-Indirect-model1)
  \end{equation}
  (where $h$ is the identity function, so that the model can be used to express risk differences)
  
  2) a model of the counterfactual values of the mediator $\mathbb{E}(M_{a} \mid L(0))=g^{-1}(a,l(0))$, where $g$ is a link function. For example with a binary mediator:
    \begin{equation}
    \mathbb{E}(M_a \mid l(0)) = g^{-1}\left[\beta_0 + \beta_A a + \beta_{L(0)} l(0) \right]
    (\#eq:MSM-Indirect-model2)
  \end{equation}
  (where $g$ is the logit function because the mediator is binary).
  
VanderWeele shows that if the function $h$ is linear in $m$ (no quadratic terms in $m$, nor transformations such as $\log(m)$ or $\sqrt{m}$, etc) and the exposure $A$ does not affect the intermediate confounder $L(1)$, then
\begin{equation*}
  \mathbb{E}(Y_{a,M_{a^*}}) = h^{-1}\left[a,g^{-1}\left(a^*,l(0)\right),l(0)\right]
\end{equation*}

Using the 2 MSMs, we can express the Natural Direct and Indirect Effects conditional on baseline confounders $L(0)$. In our example:
\begin{align*}
  \text{PNDE} \mid L(0) &= \mathbb{E}(Y_{a,M_{a^*}} \mid L(0)) - \mathbb{E}(Y_{a^*,M_{a^*}} \mid L(0)) \\
  &= \{\alpha_0 + \alpha_A a + [\alpha_M + \alpha_{AM}a] \times g^{-1}(a^*,l(0)) + \alpha_{L(0)} l(0) \} \\
  & \quad \quad - \{ \alpha_0 + \alpha_A a^*+ [\alpha_M + \alpha_{AM}a^*] \times g^{-1}(a^*,l(0)) + \alpha_{L(0)} l(0) \} \\
  &= (a - a^*) \times [\alpha_A + \alpha_{AM} \times g^{-1}(a^*,l(0))]
\end{align*}

\begin{align*}
  \text{TNIE} \mid L(0) &= \mathbb{E}(Y_{a,M_{a}} \mid L(0)) - \mathbb{E}(Y_{a,M_{a^*}} \mid L(0)) \\
  &= \{\alpha_0 + \alpha_A a + [\alpha_M + \alpha_{AM}a] \times g^{-1}(a,l(0)) + \alpha_{L(0)} l(0) \} \\
  & \quad \quad - \{ \alpha_0 + \alpha_A a+ [\alpha_M + \alpha_{AM}a] \times g^{-1}(a^*,l(0)) + \alpha_{L(0)} l(0) \} \\
  &= \left[ g^{-1}(a,l(0)) - g^{-1}(a^*,l(0)) \right](\alpha_M + \alpha_{AM} a)
\end{align*}

Marginal Natural Direct and Indirect effect can then be obtained:
\begin{align*}
  \text{PNDE} &= \sum_{l(0)} \left[ \text{PNDE} \mid L(0) = l(0) \right] \times P(L(0)=l(0)) \\
  \quad \text{TNIE} &= \sum_{l(0)} \left[ \text{TNIE} \mid L(0) = l(0) \right] \times P(L(0)=l(0))
\end{align*}

### Estimation of the 2 MSMs coefficients by IPTW for NDE and NIE
As previously, MSM coefficients can be estimated using an Inverse Probability of Treatment (IPTW) approach based on weighted regressions. 

In order to fit the 1st MSM \@ref(eq:MSM-Indirect-model1), we can use a linear regression of the (observed) outcome $Y$ on the exposure and mediator, adjusted for $L(0)$, weighted by individual stabilized weights $sw_{msm1,i}$ [@vanderweele2009]:
\begin{equation*} 
  \mathbb{E}\left(Y \mid A,M,L(0)\right) = \alpha_0 + \alpha_A a + \alpha_M m + \alpha_{AM} a \times m + \alpha_{L(0)} L(0)
\end{equation*}

where $sw_{msm1,i}$ is the product of two weights $sw_{msm1,i} = sw_{A,i} \times sw_{M,i}$, 
\begin{align*}
  sw_{A,i} =& \frac{P(A=a_i)}{P(A=a_i \mid L(0)=l(0)_i)} \quad \text{or} \quad sw_{A,i} = \frac{P(A=a_i \mid L(0)=l(0)_i)}{P(A=a_i \mid L(0)=l(0)_i)} = 1 \\
  sw_{M,i} =& \frac{P(M=m_i \mid A=a_i)}{P(M = m_i \mid A=a_i,L(0)=l(0)_i), L(1)=l(1)_i} \\
  & \quad \text{or} \quad sw_{M,i} = \frac{P(M=m_i \mid A=a_i,L(0)=l(0)_i)}{P(M = m_i \mid A=a_i,L(0)=l(0)_i), L(1)=l(1)_i} \\
\end{align*}

In order to fit the 2nd MSM \@ref(eq:MSM-Indirect-model2), we can use a logistic regression of the (observed) mediator $M$ on the exposure, adjusted for $L(0)$, weighted by individual stabilized weights $sw_{msm2,i}$:
\begin{equation*}
 \text{logit} \mathbb{E}(M \mid a,l(0)) = \beta_0 + \beta_A a + \beta_{L(0)} l(0)
\end{equation*}

\begin{equation*}
 \text{where} \quad sw_{msm2,i} = \frac{P(A=a_i)}{P(A=a_i \mid L(0)=l(0)_i)}
\end{equation*}

```{r Psi_PNDE_TNIE_MSM_IPTW, echo=TRUE, eval = FALSE}
### MSM of NDE & NIE, estimated by IPTW ----------------------------------------
## 1. Stabilized weight for the MSM1
# 1a. sw_Ai = g(A=a_i | L(0)) / g(A=a_i | L(0)) = 1
sw_Ai <- rep(1, nrow(df1_int))

# 1b. sw_Mi = g(M=m_i | A,L(0)) / g(M=m_i | A,L(0),L(1))
g.M.AL0 <- glm(M_smoking ~ A0_ace + L0_male + L0_parent_low_educ_lv,
               family = "binomial", data = df1_int)
g.Mis1.AL0 <- predict(g.M.AL0, type = "response")
sw_M.num <- rep(NA, nrow(df1_int))
sw_M.num[df1_int$M_smoking==1] <- g.Mis1.AL0[df1_int$M_smoking==1]
sw_M.num[df1_int$M_smoking==0] <- (1 - g.Mis1.AL0[df1_int$M_smoking==0])

g.M.AL0L1 <- glm(M_smoking ~ A0_ace + L0_male + L0_parent_low_educ_lv + L1,
                 family = "binomial", data = df1_int)
g.Mis1.AL0L1 <- predict(g.M.AL0L1, type = "response")
sw_M.denom <- rep(NA, nrow(df1_int))
sw_M.denom[df1_int$M_smoking==1] <- g.Mis1.AL0L1[df1_int$M_smoking==1]
sw_M.denom[df1_int$M_smoking==0] <- (1 - g.Mis1.AL0L1[df1_int$M_smoking==0])

sw_msm1 <- sw_Ai * sw_M.num / sw_M.denom

## 2. Estimate coefficients of the MSM1
MSM1 <- glm(Y_death ~ A0_ace + M_smoking + A0_ace:M_smoking +
              L0_male + L0_parent_low_educ_lv,
            weights = sw_msm1,
            family = "gaussian",
            data = df1_int)
coef(MSM1)
# (Intercept)                A0_ace             M_smoking
#  0.12033221            0.06381257            0.06691712
#     L0_male L0_parent_low_educ_lv      A0_ace:M_smoking
#  0.04671886            0.05521263            0.01652446

## 3. Stabilized weight for the MSM2
# 3a. sw_A = g(A=a_i) / g(A=a_i | L(0))
# numerator
g.A <- glm(A0_ace ~ 1, family = "binomial", data = df1_int)
g.Ais1 <- predict(g.A, type = "response")
sw_msm2.num <- rep(NA, nrow(df1_int))
sw_msm2.num[df1_int$A0_ace==1] <- g.Ais1[df1_int$A0_ace==1]
sw_msm2.num[df1_int$A0_ace==0] <- (1 - g.Ais1[df1_int$A0_ace==0])

# denominator
g.A.L0 <- glm(A0_ace ~ L0_male + L0_parent_low_educ_lv,
              family = "binomial", data = df1_int)
g.Ais1.L0 <- predict(g.A.L0, type = "response")
sw_msm2.denom <- rep(NA, nrow(df1_int))
sw_msm2.denom[df1_int$A0_ace==1] <- g.Ais1.L0[df1_int$A0_ace==1]
sw_msm2.denom[df1_int$A0_ace==0] <- (1 - g.Ais1.L0[df1_int$A0_ace==0])

# stabilized weight
sw_msm2 <- sw_msm2.num / sw_msm2.denom

## 3. Estimate coefficients of the MSM2
MSM2 <- glm(M_smoking ~ A0_ace + L0_male + L0_parent_low_educ_lv,
            weights = sw_msm2,
            family = "binomial",
            data = df1_int)
coef(MSM2)
# (Intercept)                A0_ace               L0_male L0_parent_low_educ_lv
#  -1.2723106             0.5883720             0.2566129             0.3270087

## 4. Estimate PNDE conditional on L(0), and the marginal value of PNDE
# a = 1 and a* = 0
# PNDE|L(0) = (a - a*)[alpha_A + alpha_AM.g^-1(a^*,l(0))]
g.minus1.A0 <- plogis(coef(MSM2)["(Intercept)"] + coef(MSM2)["A0_ace"] * 0 +
                      coef(MSM2)["L0_male"] * df1_int$L0_male +
                      coef(MSM2)["L0_parent_low_educ_lv"] * df1_int$L0_parent_low_educ_lv)

# PNDE conditional on L(0)
PNDE_L0 <- (1 - 0) * (coef(MSM1)["A0_ace"] +
                        coef(MSM1)["A0_ace:M_smoking"] * g.minus1.A0)
# marginal PNDE
PNDE <- mean(PNDE_L0)
# [1] 0.06850657

## 4. Estimate TNIE conditional on L(0), and the marginal value of TNIE
# TNIE|L(0) = [g^-1(a,l(0)) - g^-1(a^*,l(0))] * (alpha_M + alpha_AM * a)
g.minus1.A1 <- plogis(coef(MSM2)["(Intercept)"] + coef(MSM2)["A0_ace"] * 1 +
                        coef(MSM2)["L0_male"] * df1_int$L0_male +
                        coef(MSM2)["L0_parent_low_educ_lv"] * df1_int$L0_parent_low_educ_lv)

# TNIE conditional on L(0)
TNIE_L0 <- (g.minus1.A1 - g.minus1.A0) * (coef(MSM1)["M_smoking"] +
                                            coef(MSM1)["A0_ace:M_smoking"] * 1)
# marginal PNDE
TNIE <- mean(TNIE_L0)
# [1] 0.01096799
```

In this example, the estimation of the PNDE is 6.9\% and the estimation of the TNIE is 1.1\%. Confidence intervals can be calculated by bootstrap.
