# Appendix A: Data generating mechanisms {#appendix_a} 

The data generating mechanisms are characterized by a causal model and a statistical model that generate data given in example. 

In the first causal model, the mediator-outcome confounder $L(1)$ is not affected by the exposure. In the second causal model, the mediator-outcome confounder $L(1)$ is affected by the exposure.

## First causal model: Data generating mechanism without mediator-outcome confounder affected by the exposure

This data generating mechanism is defined by the following set of structural equations:
$$\begin{array}{lll}
P(L(0)_{male} = 1) &=& p_{L(0)_{male}}\\
P(L(0)_{soc.env} = 1) &=& p_{L(0)_{soc.env}}\\
P(A_{PM_{2.5}} = 1) &=& \beta_{A} + \beta_{male}^A \times L(0)_{male} + \beta_{soc.env}^A \times L(0)_{soc.env}\\
P(L(1) = 1) &=& p_{L(1)}\\
P(M_{smoking} = 1) &=& \beta_{M} + \beta_{male}^M \times L(0)_{male} + \beta_{soc.env}^M \times L(0)_{soc.env} + \beta_{L(1)}^M \times L(1) + \beta_{A}^M \times A_{PM_{2.5}}\\
P(Y_{death} = 1) &=& \beta_{Y} + \beta_{male}^Y \times L(0)_{male} + \beta_{soc.env}^Y \times L(0)_{soc.env} + \beta_{L(1)}^Y \times L(1)\\
                 & & + \beta_{A}^Y \times A_{PM_{2.5}} + \beta_{M}^Y \times M_{diabetes} + \beta_{A \ast M }^Y \times A_{PM_{2.5}} \times M_{diabetes}\\
\mathbb{E}(Y_{Qol} = 1) &=& \gamma_{Y} + \gamma_{male}^Y \times L(0)_{male} + \gamma_{soc.env}^Y \times L(0)_{soc.env} + \gamma_{L(1)}^Y \times L(1)\\
                 & &+ \gamma_{A}^Y \times A_{PM_{2.5}} + \gamma_{M}^Y \times M_{diabetes} + \gamma_{A \ast M }^Y \times A_{PM_{2.5}} \times M_{diabetes} + \varepsilon_Y
\end{array}$$
where $\varepsilon_Y \sim \mathcal{N}(0,\sigma_Y = 10)$.

One can set the parameters of these structural equations using the following function `param.causal.model.1()`:

```{r code_param_causal_model_1, echo=TRUE, eval=TRUE}
param.causal.model.1 <- function(A.M.interaction = NULL) {
# L0
p_L0_male <- 0.5
p_L0_soc_env <- 0.65

# A: A0_PM2.5 <- rbinom( 0.05 + 0.04 * L0_male + 0.06 * L0_soc_env ) 
b_A <- 0.05   # reference prevalence is 5%
b_male_A <- 0.04  # + 0.04 for the effect of L0_male -> A0_PM2.5
b_soc_env_A <- 0.06  # +0.06 for the effect of L0_soc_env -> A0_PM2.5

# L1: intermediate confounder between M and Y, not influenced by A
p_L1 <- 0.3

# M: M_diabetes <- rbinom( 0.2 + 0.05 * L0_male + 0.06 * L0_soc_env + 0.07 * L1 +
#                         0.1 * A0_PM2.5 ) 
b_M <- 0.2 # reference prevalence is 20%
b_male_M <- 0.05 # +0.05 for the effect of L0_male -> M_diabetes
b_soc_env_M <- 0.06 # +0.06 for the effect of L0_soc_env -> M_diabetes
b_L1_M <- 0.07 # +0.07 for the effect of L1 -> M_diabetes
b_A_M <- 0.1 # +0.10 for the effect of A0_PM2.5 -> M_diabetes

# Y binary: rbinom( 0.10 + 0.06 * L0_male + 0.04 * L0_soc_env + 0.05 * A0_PM2.5 +
#                   0.07 * L1 + 0.08 * M_diabetes +
#                   0.03 * A0_PM2.5 * M_diabetes * A.M.inter ) 
b_Y <- 0.1 # reference prevalence is 10%
b_male_Y <- 0.06 # +0.06 for the effect of L0_male -> Y
b_soc_env_Y <- 0.04 # +0.04 for the effect of L0_soc_env -> Y
b_A_Y <- 0.05 # 0.05 for the effect of A0_PM2.5 -> Y
b_L1_Y <- 0.07 # +0.07 for the effect of L1 -> Y
b_M_Y <- 0.08 # 0.08 for the effect of M_diabetes -> Y
b_AM_Y <- 0.03 # 0.03 for the interaction effect A0_PM2.5 * M_diabetes -> Y

# Y continuous: (75 - 1 * L0_male - 3 * L0_soc_env - 4 * A0_PM2.5 -3.5 * L1 - 
#                9 * M_diabetes -5 * A0_PM2.5 * M_diabetes * A.M.inter ) + 
#                rnorm(N, mean = 0, sd = 10)
mu_Y <- 75 # reference mean for QoL
c_male_Y <- -1 # -1 for the effect of L0_male -> Y
c_soc_env_Y <- -3 # -3 for the effect of L0_soc_env -> Y
c_A_Y <- -4 # -4 for the effect of A0_PM2.5 -> Y
c_L1_Y <- -3.5 # -3.5 for the effect of L1 -> Y
c_M_Y <- -9 # -9 for the effect of M_diabetes -> Y
c_AM_Y <- -5  # - 5 for the interaction effect A0_PM2.5 * M_diabetes  -> Y
sd_Y <- 10 # standard deviation of the residuals

# A*M interaction ?
A.M.inter <- A.M.interaction

coef <- c( p_L0_male = p_L0_male, p_L0_soc_env = p_L0_soc_env, 
           b_A = b_A, b_male_A = b_male_A, b_soc_env_A = b_soc_env_A, 
           p_L1 = p_L1,
           b_M = b_M, b_male_M = b_male_M, b_soc_env_M = b_soc_env_M, 
            b_L1_M = b_L1_M, b_A_M = b_A_M,
           b_Y = b_Y, b_male_Y = b_male_Y, b_soc_env_Y = b_soc_env_Y, 
            b_A_Y = b_A_Y, b_L1_Y = b_L1_Y, b_M_Y = b_M_Y, b_AM_Y = b_AM_Y,
           mu_Y = mu_Y, c_male_Y = c_male_Y, c_soc_env_Y = c_soc_env_Y, 
            c_A_Y = c_A_Y, c_L1_Y = c_L1_Y, c_M_Y = c_M_Y, c_AM_Y = c_AM_Y, 
           sd_Y = sd_Y, A.M.inter = A.M.inter)
  
return(coef)
}
```


## Second causal model: Data generating mechanism with mediator-outcome confounder affected by the exposure

This data generating mechanism is defined by the following set of structural equations:
$$\begin{array}{lll}
P(L(0)_{male} = 1) &=& p_{L(0)_{male}}\\
P(L(0)_{soc.env} = 1) &=& p_{L(0)_{soc.env}}\\
P(A_{PM_{2.5}} = 1) &=& \beta_{A} + \beta_{male}^A \times L(0)_{male} + \beta_{soc.env}^A \times L(0)_{soc.env}\\
P(L(1) = 1) &=& \beta_{L(1)} + \beta_{male}^{L(1)} \times L(0)_{male} + \beta_{soc.env}^{L(1)} \times L(0)_{soc.env} + \beta_{A}^{L(1)} \times A_{PM_{2.5}}\\
P(M_{diabetes} = 1) &=& \beta_{M} + \beta_{male}^M \times L(0)_{male} + \beta_{soc.env}^M \times L(0)_{soc.env} + \beta_{L(1)}^M \times L(1) + \beta_{A}^M \times A_{PM_{2.5}}\\
P(Y_{death} = 1) &=& \beta_{Y} + \beta_{male}^Y \times L(0)_{male} + \beta_{soc.env}^Y \times L(0)_{soc.env} + \beta_{L(1)}^Y \times L(1)\\
                 & & + \beta_{A}^Y \times A_{PM_{2.5}} + \beta_{M}^Y \times M_{diabetes} + \beta_{A \ast M }^Y \times A_{PM_{2.5}} \times M_{diabetes}\\
\mathbb{E}(Y_{Qol} = 1) &=& \gamma_{Y} + \gamma_{male}^Y \times L(0)_{male} + \gamma_{soc.env}^Y \times L(0)_{soc.env} + \gamma_{L(1)}^Y \times L(1)\\
                        & &+ \gamma_{A}^Y \times A_{PM_{2.5}} + \gamma_{M}^Y \times M_{diabetes} + \gamma_{A \ast M }^Y \times A_{PM_{2.5}} \times M_{diabetes} + \varepsilon_Y
\end{array}$$
where $\varepsilon_Y \sim \mathcal{N}(0,\sigma_Y = 10)$.

One can set the parameters of these structural equations using the following function `param.causal.model.2()`:
```{r code_param_causal_model_2, echo=TRUE, eval=TRUE}
param.causal.model.2 <- function(A.M.interaction = NULL) {
# L0
p_L0_male <- 0.5
p_L0_soc_env <- 0.65

# A: A0_PM2.5 <- rbinom( 0.05 + 0.04 * L0_male + 0.06 * L0_soc_env ) 
b_A <- 0.05   # reference prevalence is 5%
b_male_A <- 0.04  # + 0.04 for the effect of L0_male -> A0_PM2.5
b_soc_env_A <- 0.06  # +0.06 for the effect of L0_soc_env -> A0_PM2.5

# L1: L1 <- rbinom( 0.30 - 0.05 * L0_male + 0.08 * L0_soc_env + 
#                   0.2 * A0_PM2.5 ) 
b_L1 <- 0.30   # reference prevalence is 30%
b_male_L1 <- -0.05  # - 0.05 for the effect of L0_male -> L1
b_soc_env_L1 <- +0.08 # + 0.08 for the effect of L0_soc_env -> L1
b_A_L1 <- +0.2 # +0.2 for the effect of A0_PM2.5 -> L1

# M: M_diabetes <- rbinom( 0.2 + 0.05 * L0_male + 0.06 * L0_soc_env + 
#                         0.2 * L1 + 0.1 * A0_PM2.5 ) 
b_M <- 0.2 # reference prevalence is 20%
b_male_M <- 0.05 # +0.05 for the effect of L0_male -> M_diabetes
b_soc_env_M <- 0.06 # +0.06 for the effect of L0_soc_env -> M_diabetes
b_A_M <- 0.1 # +0.10 for the effect of A0_PM2.5 -> M_diabetes
b_L1_M <- 0.2 # +0.2 for the effect of L1 -> M_diabetes

# Y binary: rbinom( 0.10 + 0.06 * L0_male + 0.04 * L0_soc_env + 
#                   0.05 * A0_PM2.5 + 0.07 * L1 + 0.08 * M_diabetes +
#                   0.03 * A0_PM2.5 * M_diabetes * A.M.inter ) 
b_Y <- 0.1 # reference prevalence is 10%
b_male_Y <- 0.06 # +0.06 for the effect of L0_male -> Y
b_soc_env_Y <- 0.04 # +0.04 for the effect of L0_soc_env -> Y
b_A_Y <- 0.05 # 0.05 for the effect of A0_PM2.5 -> Y
b_L1_Y <- 0.07 # +0.07 for the effect of L1 -> Y
b_M_Y <- 0.08 # 0.08 for the effect of M_diabetes -> Y
b_AM_Y <- 0.03 # 0.03 for the interaction effect A0_PM2.5 * M_diabetes -> Y

# Y continuous: (75 - 1 * L0_male - 3 * L0_soc_env - 4 * A0_PM2.5 +
#                -3.5 * L1 - 9 * M_diabetes + 
#                -5 * A0_PM2.5 * M_diabetes * A.M.inter ) + rnorm(N, mean = 0, sd = 10)
mu_Y <- 75 # reference mean for QoL
c_male_Y <- -1 # -1 for the effect of L0_male -> Y
c_soc_env_Y <- -3 # -3 for the effect of L0_soc_env -> Y
c_A_Y <- -4 # -4 for the effect of A0_PM2.5 -> Y
c_L1_Y <- -5 # -5 for the effect of L1 -> Y
c_M_Y <- -9 # -9 for the effect of M_diabetes -> Y
c_AM_Y <- -5  # - 5 for the interaction effect A0_PM2.5 * M_diabetes  -> Y
sd_Y <- 10 # standard deviation of the residuals

# A*M interaction ?
A.M.inter <- A.M.interaction

coef <- c( p_L0_male = p_L0_male, p_L0_soc_env = p_L0_soc_env, 
           b_A = b_A, b_male_A = b_male_A, b_soc_env_A = b_soc_env_A, 
           b_L1 = b_L1, b_male_L1 = b_male_L1, b_soc_env_L1 = b_soc_env_L1, 
            b_A_L1 = b_A_L1,
           b_M = b_M, b_male_M = b_male_M, b_soc_env_M = b_soc_env_M, 
            b_L1_M = b_L1_M, b_A_M = b_A_M,
           b_Y = b_Y, b_male_Y = b_male_Y, b_soc_env_Y = b_soc_env_Y, 
            b_A_Y = b_A_Y, b_L1_Y = b_L1_Y, b_M_Y = b_M_Y, b_AM_Y = b_AM_Y,
           mu_Y = mu_Y, c_male_Y = c_male_Y, c_soc_env_Y = c_soc_env_Y, 
            c_A_Y = c_A_Y, c_L1_Y = c_L1_Y, c_M_Y = c_M_Y, c_AM_Y = c_AM_Y, 
              sd_Y = sd_Y, A.M.inter = A.M.inter)
  
  return(coef)
}
```


## Simulation of the four data sets used in examples

### Data sets generated from the causal model 1

The following function `gen.data.causal.model.1` can be used to simulate data sets using the parameters defined previously in the `param.causal.model.1` function.

```{r gen_causal_model_1_dataset, echo=TRUE, eval = TRUE}
gen.data.causal.model.1 <- function(N, A.M.inter) { # input parameters are the 
  #   sample size N and the presence of A*M interaction with A.M.inter = 0 or 1
  
  b <- param.causal.model.1(A.M.interaction = A.M.inter)
    
  # baseline confounders: parent's educational level=L0_soc_env & sex=L0_male
  L0_male <- rbinom(N, size = 1, prob = b["p_L0_male"]) 
  L0_soc_env <- rbinom(N, size = 1, prob = b["p_L0_soc_env"])  
  
  # exposure: A0_PM2.5
  A0_PM2.5 <- rbinom(N, size = 1, prob =  b["b_A"] + 
                     b["b_male_A"] * L0_male + 
                     b["b_soc_env_A"] * L0_soc_env ) 
  
  # intermediate confounder between M_diabetes and Y, not affected by A0 L1
  L1 <- rbinom(N, size = 1, prob = b["p_L1"])
  
  # mediator: M_diabetes
  M_diabetes <- rbinom(N, size = 1, prob = b["b_M"] + 
                        b["b_male_M"] * L0_male + 
                        b["b_soc_env_M"] * L0_soc_env + 
                        b["b_A_M"] * A0_PM2.5 +
                        b["b_L1_M"] * L1) 

  # Y_death 
  Y_death <- rbinom(N, size = 1, prob = b["b_Y"] + 
                      b["b_male_Y"] * L0_male + 
                      b["b_soc_env_Y"] * L0_soc_env + 
                      b["b_A_Y"] * A0_PM2.5 + 
                      b["b_L1_Y"] * L1 +
                      b["b_M_Y"] * M_diabetes +
                      b["b_AM_Y"] * A0_PM2.5 * M_diabetes * A.M.inter ) 
  
  # Y_qol 
  Y_qol <- ( b["mu_Y"] + 
               b["c_male_Y"] * L0_male + 
               b["c_soc_env_Y"] * L0_soc_env +
               b["c_A_Y"] * A0_PM2.5 +
               b["c_L1_Y"] * L1 +
               b["c_M_Y"] * M_diabetes + 
               b["c_AM_Y"] * A0_PM2.5 * M_diabetes * A.M.inter ) + 
    rnorm(N, mean = 0, sd = b["sd_Y"])
  
  # data.frame
  data.sim <- data.frame(L0_male, L0_soc_env, A0_PM2.5, L1, M_diabetes, 
                         Y_death, Y_qol)

  return( data.sim )
}
```

Applying a sample size N=10000, we generate the `df1.csv` and `df1_int.csv` data sets.

```{r simulate_data_1, echo=TRUE, eval = TRUE}
set.seed(1234)
df1 <- gen.data.causal.model.1(N=10000, A.M.inter=0)
write.csv(df1, file = "data/df1.csv", row.names = FALSE)

set.seed(1234)
df1_int <- gen.data.causal.model.1(N=10000, A.M.inter=1)
write.csv(df1_int, file = "data/df1_int.csv", row.names = FALSE)
```

```{r show_data_df1, echo=TRUE, eval = TRUE}
head(df1)
```

```{r show_data_df1_int, echo=TRUE, eval = TRUE}
head(df1_int)
```


### Data sets generated from the causal model 2

The following function `gen.data.causal.model.2` can be used to simulate data sets using the parameters defined previously in the `param.causal.model.2` function.

```{r gen_causal_model_2_dataset, echo=TRUE, eval=TRUE}
gen.data.causal.model.2 <- function(N, A.M.inter) { # input parameters are the 
  #   sample size N and the presence of A*M interaction with A.M.inter = 0 or 1
  
  b <- param.causal.model.2(A.M.interaction = A.M.inter)
    
  # baseline confounders: parent's educational level=L0_soc_env & sex=L0_male
  L0_male <- rbinom(N, size = 1, prob = b["p_L0_male"]) 
  L0_soc_env <- rbinom(N, size = 1, prob = b["p_L0_soc_env"])  
  
  # exposure: A0_PM2.5
  A0_PM2.5 <- rbinom(N, size = 1, prob =  b["b_A"] + 
                     b["b_male_A"] * L0_male + 
                     b["b_soc_env_A"] * L0_soc_env ) 
  
  # intermediate confounder between M_diabetes and Y, 
  L1 <- rbinom(N, size = 1, prob = b["b_L1"] +
                 b["b_male_L1"] * L0_male +
                 b["b_soc_env_L1"] * L0_soc_env + 
                 b["b_A_L1"]* A0_PM2.5)
  
  # mediator: M_diabetes
  M_diabetes <- rbinom(N, size = 1, prob = b["b_M"] + 
                        b["b_male_M"] * L0_male + 
                        b["b_soc_env_M"] * L0_soc_env + 
                        b["b_A_M"] * A0_PM2.5 +
                        b["b_L1_M"] * L1) 

  # Y_death 
  Y_death <- rbinom(N, size = 1, prob = b["b_Y"] + 
                      b["b_male_Y"] * L0_male + 
                      b["b_soc_env_Y"] * L0_soc_env + 
                      b["b_A_Y"] * A0_PM2.5 + 
                      b["b_L1_Y"] * L1 +
                      b["b_M_Y"] * M_diabetes +
                      b["b_AM_Y"] * A0_PM2.5 * M_diabetes * A.M.inter ) 
  
  # Y_qol 
  Y_qol <- ( b["mu_Y"] + 
               b["c_male_Y"] * L0_male + 
               b["c_soc_env_Y"] * L0_soc_env +
               b["c_A_Y"] * A0_PM2.5 +
               b["c_L1_Y"] * L1 +
               b["c_M_Y"] * M_diabetes + 
               b["c_AM_Y"] * A0_PM2.5 * M_diabetes * A.M.inter ) + 
    rnorm(N, mean = 0, sd = b["sd_Y"])
  
  # data.frame
  data.sim <- data.frame(L0_male, L0_soc_env, A0_PM2.5, L1, M_diabetes, 
                         Y_death, Y_qol)

  return( data.sim )
}
```

Applying a sample size N=10000, we generate the `df2.csv` and `df2_int.csv` data sets.

```{r simulate_data_2, echo=TRUE, eval = TRUE}
set.seed(1234)
df2 <- gen.data.causal.model.2(N=10000, A.M.inter=0)
write.csv(df2, file = "data/df2.csv", row.names = FALSE)

set.seed(1234)
df2_int <- gen.data.causal.model.2(N=10000, A.M.inter=1)
write.csv(df2_int, file = "data/df2_int.csv", row.names = FALSE)
```

```{r show_data_df2, echo=TRUE, eval = TRUE}
tail(df2)
```

```{r show_data_df2_int, echo=TRUE, eval = TRUE}
tail(df2_int)
```

